# Introduction to Statistics

## Statistics

Researchers in the life sciences analyze **data** that exhibit some **variability**. For example, patients given the same treatment may respond differently; PCR (Polymerase Chain Reaction) prepared identically may have different results due to contamination, thermal damage, etc. The challenge to the life scientist is to discern the **signal** from the **noise**.

**Statistics** is the science of understanding data (collection, analysis and presentation) and of making decisions in the face of variability and uncertainty. Statistics is a core component of **data science**.

![Data Science, credit: Lee et al. 2018 ](/Figures/DataScience.ppm){fig-align="center" width="50%"}

::: {.callout-note icon=false}
## Example: Cats in New York
An article presented results on the number of injuries of cats that had plummeted from buildings in New York City, according to the number of floors they had fallen. The damage caused by such falls was dubbed feline high-rise syndrome (FHRS).
:::

![Raining cats](/Figures/cats){fig-align="center" width="90%"}

What can we conclude?

High-quality samples are essential for sound science, even in the age of AI!


Crucially, statistics gives us tools to **quantify the uncertainty** of certain quantities of interest.

Our ability to obtain reliable measures of population quantities of interest, and to assess uncertainty, depends on how we **sample** populations. It is often at this early step in an investigation that the fate of a study is sealed for better or worse.


![Statistics](/Figures/pop_cats){fig-align="center" width="95%"}


In most situations, we do not observe the whole population. Instead we obtain a sample. What we need to ask ourselves is what do we want to do with the data? If our goal is to describe the sample data, we generate **descriptive statistics** and perform an **Exploratory Data Analysis (EDA)**. 


![EDA https://r-graph-gallery.com/](/Figures/EDA){fig-align="center" width="95%"}


If instead, we want to make statements about the whole population, then apart from EDA, we do **statistical inference**. 


![Statistical Inference](/Figures/inference){fig-align="center" width="95%"}

## Descriptive statistics

Your (tabular) data is a collection of variables and observations. Typically one observation per row, one variable per column and filled in with values. A **variable** is a characteristic or quantity that you can you measure. A **value** is the state of the variable when you measure it. An **observation** is a set of measurements. 

Variables can be **categorical** or **numerical**. Categorical variables tell us membership or attribute. Examples: blood type (A,B,AB, O), survival (alive or dead), language, etc. Numerical variables record the amount of something, for example body temperature, age, number of friends. Numerical values can be discrete (number of friends) or continuous (body temperature).   



The most common languages used for analyzing data are SAS, R (and Bioconductor) and Python. In this course, we will be using R with datasets in the form of **data frames** or  **tibbles**.

```{r,echo=FALSE}
library(qrcode)
qr <- qrcode::qr_code("https://docs.google.com/forms/d/e/1FAIpQLSecc961cTxMxEiZTbzGC3luAM6xE8atUWFM4-_mjCDXZBmhWA/viewform?usp=publish-editor")
generate_svg(qr, filename = "Figures/poll.svg")
```
  

![poll](/Figures/poll.svg){fig-align="center" width="40%"}
We can now read our dataset (poll results)
```{r}
library(readr)

url <- "https://github.com/JuliaPalacios/stats141_2026/data/file1.csv"
survey <- read_csv(url) 
survey

```


## Types of studies

Researchers gather information to make inferences about nature in a variety of settings. Much of statistics deals with **data analysis**, but statistical considerations often play a key role in the planning and design of scientific investigation. 

Data in biology are obtained from either an **experimental study** or an **observational study**. In an experimental study, the researcher assigns different treatment groups or values of an explanatory variable randomly to the individual units of study. In an observational study, on the other hand, the researcher has no control over which units fall into which groups.


         
::: {.callout-note icon=false}

## Example: Marriage and Life expectancy
A study found that people who were married are more likely to live longer than single. However, from an **observational** study such as this we don't know whether marriage increases life expectancy or whether persons who are healthier are more likely to get married. 
:::

      
::: {.callout-note icon=false}

## Example: Aids as confounder
Some research has suggested that there is a biological basis for sexual orientation. One such study involved measuring a section of the brain for 40 homosexual men, 40 heterosexual men, and 40 heterosexual women. The research found that the brain section tends to be larger in homosexual men and heterosexual women than in heterosexual men.  
:::

```{r, echo=FALSE}
library(knitr)
set.seed(123)

n <- 40
brain <- data.frame(
  group = factor(rep(c("Homosexual men", "Heterosexual men", "Heterosexual women"),
                     each = n)),
  brain_area_mm2 = c(
    rnorm(n, mean = 15.1, sd = 1.0),
    rnorm(n, mean = 11.2, sd = 1.0),
    rnorm(n, mean = 12.0, sd = 1.0)
  )
)

# check group means (will be close, not exact)
mean_table<-aggregate(brain_area_mm2 ~ group, data = brain, mean)
names(mean_table)[2] <- "Brain Area"
kable(mean_table, digits = 2, caption = "Average brain area (mm^2) by group")
x <- jitter(as.numeric(brain$group), amount = 0.12)
cols <- rep("steelblue", nrow(brain))

# indices within each group to color red
i_hom <- which(brain$group == "Homosexual men")[sample(1:40,5)]
i_hetm <- which(brain$group == "Heterosexual men")[sample(1:40,32)]
i_hetw <- which(brain$group == "Heterosexual women")[sample(1:40,39)]

cols[c(i_hom, i_hetm, i_hetw)] <- "red"

plot(x, brain$brain_area_mm2,
     xaxt = "n",
     xlab = "Group", ylab = expression("Brain area (mm"^2*")"),
     pch = 16, col = cols)

axis(1, at = 1:3, labels = levels(brain$group))
legend("topleft",
       legend = c("AIDS", "No AIDS"),
       col = c("red", "black"),
       pch = 16, bty = "n")
```



The previous two examples are based on **observational studies**. In an observational study the researcher systematically collects data from subjects, but only as an observer and not as someone who is controling the conditions. However, observational studies can be misleading due to **confounding variables**. The context in which the data arose is important and tell us to be on the alert for the effects that other factors, such as the impact of AIDS, may have on the size of the section of the brain. 

**Confounders** are variables, other than the treatment, which are responsible for causing (at least part of) the observed difference in the outcome between the various levels of the treatment.

**How can we mitigate confounding?** Create treatment and control groups that are as similar as possible except for the treatment level and we can anticipate and prevent both other known and unknown explanations for the observed difference in outcomes between the treated and control groups.


::: {.callout-note icon=false}

## Example: Effect of drug on cats
In part of one study, a new drug for lowering cholesterol was given to seven male and seven female cats at doses of $2 mg/kg$ and $6 mg/kg$. The two doses were assigned at random to the seven cats within each sex and cholesterol levels were measured after some time.
:::

::: {.columns}
::: {.column width="50%"}

```{r,echo=FALSE}
# Packages
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
})


set.seed(7)

n <- 5
sd <- 5

spec <- tibble::tibble(
  Dose = c("2mg","2mg","6mg","6mg"),
  Sex  = c("Male","Female","Male","Female"),
  target_mean = c(140, 130, 120, 150)
)

df <- spec |>
  group_by(Dose, Sex) |>
  do({
    mu <- .$target_mean[1]
    vals <- rnorm(n, mean = mu, sd = sd)
    tibble::tibble(Value = vals, CatNum = seq_len(n))
  }) |>
  ungroup() |>
  mutate(CatID = paste0(Dose, "_", substr(Sex, 1, 1), CatNum)) |>
  select(CatID, Sex, Dose, Value)

# Nicer summary table
sum_tbl <- df |>
  group_by(Sex, Dose) |>
  summarise(
    mean = mean(Value),
    .groups = "drop"
  ) 

# Table (HTML) with kableExtra
sum_tbl |>
  tidyr::pivot_wider(names_from = Dose, values_from = mean) |>
  kbl(
    caption = "Summary by Sex and Dose",
    col.names = c("Sex","Mean (2mg)", "Mean (6mg)"),
    align = "lcc"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) |>
  column_spec(1, bold = TRUE)
```
:::
:::{.column width="50%"}
```{r,echo=FALSE}
ggplot(df, aes(x = Sex, y = Value, color = Dose)) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.5),
    alpha = 0.85, size = 2.6
  ) +
  stat_summary(
    fun = mean, geom = "point",
    position = position_dodge(width = 0.5),
    shape = 18, size = 5, color = "black"
  ) +
  labs(title = "Outcome by sex, split by dose", y = "Outcome") +
  theme_minimal(base_size = 16) +   # increases all text
  theme(
    plot.title = element_text(size = 18),
    axis.title = element_text(size = 16),
    axis.text  = element_text(size = 14),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13),
    legend.position = "top"
  )

```
:::

The **design of this experiment** allows for the investigation of the **interaction** between two factors: sex of the cat and dose. That is, for females, the effect of increasing the dose was positive in lowering cholesterol, but for males the
effect of increasing the dose was negative.

By **randomly assigning** treatments (drug doses) to subjects (dogs), we can get around the problem of confounding that complicates observational studies and limits the conclusions. **Randomized experiments** are considered the gold standard in scientific investigation, but they can also be difficult.

Often, human subjects in experiments are given a **placebo**. Of course, if a placebo control is used, then the subjects must not be told which group they are in, that is, the experiment is **blinded**. The purpose of blinding is to minimize the extent to which subject expectations influence the results of the experiment. If subjects exhibit a psychological reaction to getting a medication, that placebo response will tend to balance out between the two groups, so that any difference between the groups can be attributed to the effect of the active treatment.

![Placebo](/Figures/placebo){fig-align="center" width="50%"}

How about the person evaluating the responses? This person can be subject to bias influencing the observation process itself. An experiment in which both the subjects and the persons making the evaluations of the response are blinded is called **double-blinded**. 


Randomized allocation in medical settings may not be ethical. One possibility is to use **historical controls** (i.e. previous patients with the same illnes who were treated with another therapy). The challenge is that overall characteristics of the patients may change with time. A possible solution is to do some statistical adjustment to control for differences in characteristics (for example age).

**Random Sampling**


**Exploratory Data Analysis (EDA)**

**Confirmatory Data Analysis (CDA)**

See @knuth84 for additional discussion of literate programming.

```{r}
1 + 1
```
