---
title: "Linear regression and correlation 2"
format: html
bibliography: references.bib
---

Recall the example:

::: {.callout-note icon="false"}
**Example: Cat's Heart Weight**\
Linear regression of cat's heart weight on body weight, using the dataset $cats$ from the MASS library in R. 
:::

```{r}
library(MASS)
model1 <- lm(Hwt ~ Bwt, data = cats)
model1

```

  

### Estimating the coefficients

The coefficients $\beta_{0}$ and $\beta_{1}$ in the linear model are unknown. The goal is therefore to obtain estimates $\hat{\beta}_{0}$ and $\hat{\beta}_{1}$ such that the linear model fits the $n = 144$ data points as closely as possible. 

There are several ways to measure *closeness*, however the most common approach involves minimizing the *least squares* criterion.

In least squares, we calculate the difference between the observed value $y_{i}$ and the predicted value $\hat{y}_{i}$ obtained from the fitted model: $\hat{y}_{i} = \hat{\beta}_{0} + \hat{\beta}_{1}x_{i}$. 

That is, we minimize the mean squared error:
$$
\min \text{MSE}=\min \sum^{n}_{i=1}\frac{(y_{i}-\hat{y}_{i})^2}{n}
$$

Let
$$
e_{i} = y_{i} - \hat{y}_{i}
$$ 
represents the *i-th residual* for observation $i$ then, equivalently our goal is to minimize the *residual sum of squares*:
  
  $$RSS = e^{2}_{1} + e^{2}_{2} + \dots + e^{2}_{n}$$ 



We illustrate these residuals for the $cats$ sample in the plot below, shown as vertical lines from each observed data point (in red) to the fitted line.

```{r}
#| echo: false
# Fitted values
library(MASS)
model1 <- lm(Hwt ~ Bwt, data = cats)
model1

fitted_vals <- fitted(model1)

plot(cats$Bwt, cats$Hwt,
     col = "red", pch = 19,
     xlab = "Body Weight (kg)",
     ylab = "Heart Weight (g)",
     main = "Simple Linear Regression: Heart Weight vs Body Weight")

# Fitted line
abline(model1, col = "black", lwd = 2)

# Vertical lines from each point to the fitted line (residuals)
segments(cats$Bwt, cats$Hwt,
         cats$Bwt, fitted_vals,
         col = "lightgrey", lwd = 0.8)

# Extract coefficients
b0 <- round(coef(model1)[1], 3)
b1 <- round(coef(model1)[2], 3)

# Add label to plot
legend("topleft",
       legend = bquote(hat(Hwt) == .(b0) + .(b1) * Bwt),
       bty = "n",
       cex = 1.1)
```


    
The *least squares coefficient estimates* $\hat{\beta}_{0}$ and $\hat{\beta}_{1}$ for simple linear regression are obtained by minimizing the RSS, which after some calculus yields:
      
$$\begin{align}
    \hat{\beta}_{1} &= \frac{\sum_{i=1}^{n}(x_{i}-\bar{x})(y_{i}-\bar{y})}{\sum_{i=1}^{n}(x_{i}-\bar{x})^2}=r \left(\frac{s_{y}}{s_{x}} \right), \\
    \hat{\beta}_{0} &= \bar{y} - \hat{\beta}_{1}\bar{x},
    \end{align}
$$ {#eq-4}
      

where $\bar{y}\equiv \frac{1}{n}\sum_{i=1}^{n}y_{i}$ and $\bar{x}\equiv \frac{1}{n}\sum_{i=1}^{n}x_{i}$ are the sample means.
      
### Assessing the accuracy of the coefficient estimates
      
A natural question about the linear regression model is how accurate the coefficient estimates are — that is, how far $\hat{\beta}_{0}$ and $\hat{\beta}_{1}$ are from the true values $\beta_{0}$ and $\beta_{1}$. To answer this question, we compute the *standard errors* associated with $\hat{\beta}_{0}$ and $\hat{\beta}_{1}$ using the following formulas:
        
$$SE(\hat{\beta}_{0})^{2}=\sigma^{2}\left[\frac{1}{n}+\frac{\bar{x}^{2}}{\sum_{i=1}^{n}(x_{i}-\bar{x})^{2}}\right], \quad SE(\hat{\beta}_{1})^{2}=\frac{\sigma^{2}}{\sum_{i=1}^{n}(x_{i}-\bar{x})^{2}}$$ {#eq-5}
          
where $\sigma^{2}=Var(\epsilon)$ is the variance of the error terms, which is unknown but can be estimated empirically via the RSE defined as follows:
            
$$
RSE=\sqrt{\frac{RSS}{n-2}}
$$
              
Strictly, these equations are valid only when the errors $\epsilon_{i}$ for each observation have constant variance $\sigma^{2}$ and are uncorrelated, but in practice they provide a good approximation. 
            
Notice that unlike $e_{i}$, which are the observed residuals obtained from the fitted model, $\epsilon_{i}$ are the underlying population errors that arise when approximating the relationship between $X$ and $Y$ through a linear model, and are therefore unobserved.
            
Once we obtain the standard errors of the estimates, we can construct *confidence intervals*. A 95% confidence interval is defined as a range of values such that, with 95% probability, the range will contain the true unknown value of the parameter. Confidence intervals are constructed from the data and have the property that if we take repeated samples and construct a confidence interval for each sample, 95% of those intervals will contain the true unknown value of the parameter. For the linear regression model, the 95% confidence intervals for $\beta_{0}$ and $\beta_{1}$ are approximated as:
              
$$\left[\hat{\beta}_{0}-2\cdot SE(\hat{\beta}_{0}),\ \hat{\beta}_{0}+2\cdot SE(\hat{\beta}_{0})\right]$$ {#eq-6} $$\left[\hat{\beta}_{1}-2\cdot SE(\hat{\beta}_{1}),\ \hat{\beta}_{1}+2\cdot SE(\hat{\beta}_{1})\right]$$ {#eq-7}
                
In the previous example, the standard errors for the intercept and slope were 0.6923 and 0.2503, respectively. The associated confidence intervals are therefore:
                  
 ```{r}
 confint(model1)
 ```
 
We are thus 95% confident that for every 1 kg increase in body weight, the true average increase in heart weight is between 3.54 and 4.53 grams.
                
Standard errors are also useful for testing the most common hypothesis in regression — that there is no relationship between $X$ and $Y$ (i.e., that a cat's body weight is not associated with its heart weight). Formally, the *null hypothesis* states that there is no relationship between $X$ and $Y$, versus the *alternative hypothesis* that there is some relationship between $X$ and $Y$. Mathematically, this corresponds to testing:

$$\begin{align}
H_{0} &: \beta_{1} = 0 \\
&\text{versus} \\
H_{a} &: \beta_{1} \neq 0
\end{align}$$ {#eq-8}

To test the null hypothesis, we need to determine how far $\hat{\beta}_{1}$ is from zero to be confident that $\beta_{1}$ is nonzero. This depends on the accuracy of the estimate, given by $SE(\hat{\beta}_{1})$. The smaller $SE(\hat{\beta}_{1})$ is, the more likely that $\beta_{1}$ differs from zero; conversely, the larger the standard error, the larger $\hat{\beta}_{1}$ needs to be in order to reject the null hypothesis. In practice, we rely on a *t-statistic*, given by

$$t=\frac{\hat{\beta}_{1}-0}{SE(\hat{\beta}_{1})},$$ {#eq-9}

which measures how many standard deviations $\hat{\beta}_{1}$ is from zero. If there is no relationship between $X$ and $Y$, we expect $t$ to follow a $t$-distribution with $n-2$ degrees of freedom. The larger the observed value of $t$ relative to the theoretical $t$-distribution, the lower the probability that $\beta_{1} = 0$. This probability is called the *p-value*, which in our example was practically zero ($<2\times10^{-16}$), providing strong evidence that $\hat{\beta}_{1} \neq 0$, or equivalently that it is *statistically significant*. A small *p-value* indicates that it is unlikely to observe such a substantial association between the predictor and the response by chance alone. We therefore reject the null hypothesis when the *p-value* is sufficiently small — typically below 0.05 or 0.01.


