---
title: "Analyzing proportions and Goodness-of-fit "
format: html
bibliography: references.bib
---

**Example**: A species of crocodile nests in sandy riverbanks, and the sex of the hatchlings is determined by the temperature of the surrounding sand during incubation. Historically, typical nest depths and shading produced an approximately 50:50 male-female sex ratio. Recently, hotter seasons and reduced vegetation cover are suspected to be warming the sand and shifting the sex ratio.

Researchers recorded the sex of 180 hatchlings emerging from sandy nests along a river in northern Australia. In their sample, 54 hatchlings were male and 126 were female.

What is the estimate of the proportion of female crocodiles in this population?

The sample proportion:

$$
\hat{p}=\frac{x}{n}=\frac{126}{180}=0.7
$$

How precise is this estimate?

We can measure precision using the standard error of the estimated proportion:

$$
SE(\hat{p})=\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=0.034
$$

What is the "exact" 95% confidence interval of the population proportion?

In our example:

```{r}

lower<-qbinom(.025,180,.7)/180
upper<-qbinom(.975,180,.7)/180
print(paste("95% CI is: (",lower,upper,")"))

```

An alternative approximate CI is

$$
\hat{p}\pm 1.96 SE(\hat{p})
$$ That is:

```{r}
phat<-126/180
se<-sqrt(phat*(1-phat)/180)
lower<-phat-1.96*se
upper<-phat+1.96*se
print(paste("95% CI is: (",lower,upper,")"))

```

where $1.96$ is the 0.975 quantile of the standard normal distribution.

However a common approximation is the Agresti-Coull approximation that estimates the population proportion by

$$
\tilde{p}=\frac{x+2}{n+4}=\frac{128}{184}=0.696
$$

with

$$
SE_{\tilde{p}}=\sqrt{\frac{\tilde{p}(1-\tilde{p})}{n+4}}=0.033
$$ and 95% confidence interval:
$$
\hat{p}\pm 1.96SE_\tilde{p}
$$

```{r}
n<-180
tilde_p<-128/184
se<-sqrt(tilde_p*(1-tilde_p)/(n+4))
print(paste("95% CI is: (",tilde_p-1.96*se,tilde_p+1.96*se,")"))

```

## Binomial test

Example 2: A study of 25 genes involved in spermatogenesis found their locations in the mouse genome. The study was carried out to test a prediction of evolutionary theory that such genes should occur disproportionately often on the X chromosome. As it turned out, 10 out of 25 spermatogenesis genes (40%) were on the X chromosome. If genes for spermatogenesis occurred "randomly" throughout the genome, then we would expect only 6.1% of them to fall on the X chromosome, because the X chromosome contains 6.1% of the genes in the genome. Do the results support the hypothesis that spermatogenesis genes occur preferentially on the x chromosome?

$H_{0}: p=0.061$, the probability that a spermatogenesis gene falls on the X chromosome is $.061$.

$H_{A}: p \neq 0.061$

```{r}

pval<-2*(1-pbinom(9,25,.061))
pval
```

This p-value is below the significance level of $\alpha=.05$, therefore, we reject the null hypothesis and conclude that there is a disproprotionate number of spermatogenesis genes on the X chromosome.

## Goodness of fit test

The binomial test just introduced is an example of goodness-of-fit test. A **goodness-of-fit test** is a method for comparing an observed frequency distribution with the frequency distribution that would be expected under a simple probability model governing the occurence of different outcomes. In particular, we will use a test statistic called $\chi^2$ (chi squared) calculated as follows:

$$
\chi^{2}=\sum_{i}\frac{(\text{Observed}_{i}-\text{Expected}_{i})^2}{\text{Expected}_{i}}
$$

Where $\text{Observed}_{i}$ is the frequency of individuals observed in the $i$-th category and $\text{Expected}_{i}$ is the frequency expected in that category under the null hypothesis.

::: {.callout-note icon="false"}
## Example: the proportional model
We would expect that babies should be born at the same frequency on all seven days of the week. To test this, we obtained the number of babies born on each day of the week in a random sample of 180 births from 2016. 
:::

The observe data
```{r}
#| warning: false
day<-c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")
obs<-c(14,26,34,21,27,38,20)
obsp<-obs/sum(obs)
df <- data.frame(day = factor(day, levels = day), obs = obs)
library(ggplot2)

ggplot(df, aes(x = day, y = obs)) +
  geom_col(fill = "steelblue") +
  labs(x = "Day", y = "Observations", title = "Observations by Day") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Under the proportional model (null hypothesis), the number of births on Monday is proportional to the number of Mondays in 2016. 
$H_{0}:$ The frequency of births on each day of the week is proportional to the number of times that day occurs in the year


$H_{A}:$ The frequency of births on each day of the week is NOT proportional to the number of times that day occurs in the year

The expected proportion under the null is then:
```{r}
exp<-c(52,52,52,52,52,53,53)*180/366

chi_square<-sum ((obs-exp)^2 / exp)
chi_square
```

The null distribution of the $\chi^{2}$ statistic is well approximated by the theoretical $\chi^2$ distribution with degrees of freedom calculated as follows:

$$
df=\text{Num. of categories} - 1 = \text{Num. of parameters estimated from the data}
$$

```{r}
# Chi-square distributions with df = 10 and df = 20

df_vals <- c(2, 10)

# x-range covering essentially all mass for both
x <- seq(0, qchisq(0.999, df = max(df_vals)), length.out = 2000)

library(ggplot2)

dat <- data.frame(
  x = rep(x, times = length(df_vals)),
  df = factor(rep(df_vals, each = length(x))),
  density = c(dchisq(x, df = 2), dchisq(x, df = 10))
)

ggplot(dat, aes(x = x, y = density, color = df)) +
  geom_line(linewidth = 1) +
  labs(
    title = "Chi-square density",
    x = "x",
    y = "Density",
    color = "Degrees of freedom"
  ) +
  theme_minimal()
```

**Calculating the p-value**

For the $\chi^2$ goodness-of-fit test, the p-value is the probability of getting a $\chi^2$ value **greater** than the observed test value calculated from the data (why?).

```{r}
pval<-1-pchisq(chi_square,df=6)
pval
```
At the standard significance level of $\alpha=0.05$, such a small p-value leads us to reject the null hypothesis.

Assumptions:

-   Individuals in the data set are a random sample from the whole population

-   None of the categories should have an expected frequency less than one

-   No more than 20% of the categories should have expected frequencies less than five

If one of these conditions is not met, then we can combine some of the categories with small expected frequencies.


## Goodness-of-fit when there are 2 categories

::: {.callout-note icon="false"}
## Example: migraine headache
Patients who suffered from moderate to severe migraine headache took part in a double-blind clinical trial to assess an experimental surgery. A group of 75 patients were randomly assigned to receive either the real surgery on migraine trigger sites (n=49) or a sham surgery (n=26) in which an incision was made but no further procedure was performed. The surgeons hoped that patients would experience a substantial reduction in migraine headaches. 
:::

```{r}
#| warning: false

# Contingency table: rows = (Success, No success), cols = (Real, Sham)
tab <- matrix(
  c(41, 8,
    15, 11),
  nrow = 2, byrow = FALSE,
  dimnames = list(
    Outcome = c("Success", "No success"),
    Group   = c("Real", "Sham")
  )
)

#tab
addmargins(tab)  # show row/column totals
```
The focus of interest in a **contingency table** is the dependence or association between the column variable and the row variable (between treatment and response). Under the null hypothesis, there is no such association.

$H_{0}:P(\text{Success}\mid \text{Real})=P(\text{Success}\mid \text{Sham})$

$H_{A}:P(\text{Success}\mid \text{Real}) \neq P(\text{Success}\mid \text{Sham})$

Under the null, the estimated success rate is $56/75=0.746$ and so, th expected values are:
```{r}
exp<-matrix(0,nrow=2,ncol=2)
exp[1,]<-c(49*56/75,26*56/75)
exp[2,]<-c(49,26)-exp[1,]
exp

```

Then, the chi-square statistic is

```{r}
chi<-sum(((exp-tab)^2)/exp)
chi

pval<-1-pchisq(chi,df=1)
pval
```
Thus, we reject the null and find that the data provide sufficient evidence to conclude that the real surgery is better than the sham surgery for reducing migraine headache.

## Independence and association

::: {.callout-note icon="false"}
## Example: hair color and eye color
To study the relationship between hair color and eye color in a German population, an anthropologist observed a sample of 6800 men, with the following results:  
:::

```{r}
tab <- matrix(
  c(726, 3129,
    131, 2814),
  nrow = 2, byrow = FALSE,
  dimnames = list(
    Eye_Color = c("Dark", "Light"),
    Hair_Color   = c("Dark", "Light")
  )
)

#tab
addmargins(tab)

```
$H_{0}:$ Eye color is independent of hair color

$H_{A}:$ Eye color and is not independent of hair color

```{r}
exp<-matrix(0,nrow=2,ncol=2)
exp[1,]<-c(3855*857/6800,2945*857/6800)
exp[2,]<-c(3855,2945)-exp[1,]
exp
```

```{r}
chi<-sum(((exp-tab)^2)/exp)
chi

pval<-1-pchisq(chi,df=1)
pval

```


We reject the null hypothesis.