---
title: "WMW test and Paired t-test"
format: html
bibliography: references.bib
---

## The Wilcoxon-Mann-Whitney (WMW) Test

The WMW test is used to compare two independent samples like the t-test, but unlike the t-test, the WMW test is valid even when the population distributions are not normal.

$H_{0}:$ The population distributions $Y_{1}$ and $Y_{2}$ are the same.

$H_{A}:$ The population distribution of $Y_{1}$ is shifted from the population distribution of $Y_{2}$.

The WMW test statistic is denoted $U_{s}$ and measures the degree of separation or shift between the two samples.

The test procedure:

1.  Arrange the observations in increasing order

2.  Create two counts $K_{1}$ and $K_{2}$ as follows:

    1.  For $K_{1}$, count 1 for all observations in sample 2 which are less than each observation in sample 1. Count 0.5 for all observations in sample 2 which are the same value as an observation in sample 1.

    2.  For $K_{2}$ same thing but in reverse.

    3.  $U_{s}=\max\{K_{1},K_{2}\}$

Example: Soil respiration is a measure of microbial activity in soil, which affects plant growth. In one study, soil cores were taken from two locations in a forest: (1) under an opening in the forest canopy (the “gap” location) and (2) at a nearby area under heavy tree growth (the “growth” location). The amount of carbon dioxide given off by each soil core was measured (in mol CO2/g soil/hr).

$H_{0}:$ The gap and growth areas do not differ with respect to soil respiration.

$H_{A} (nondirectional):$ The distribution of soil respiration rates tends to be higher in one of the two populations.

$H_{A} (directional):$ Soil respiration rates tend to be greater in the growth area than they are in the gap area.

```{r}
#| warning: false
y1<-c(6,13,14,15,16,18,22,29)
y2<-c(17,20,22,64,170,190,315)

dat<-data.frame(
  group = factor(c(rep("Gap data",length(y2)),rep("Growth data",length(y1)))),
  values = c(y2,y1)
)

library(ggplot2)

ggplot(dat, aes(x = group, y = values, fill = group)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.8, size = 2) +
  labs(
    x = NULL,
    y = "carbon dioxide given",
    title = "mole CO2/g soil/hr"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

w <- wilcox.test(values ~ group, data = dat, exact = FALSE)
w
#one sided
w2 <- wilcox.test(values ~ group, data = dat, alternative="greater",exact = FALSE)
w2

```

-   The required assumption in the WMW test is that the samples must be random samples from their populations, independently drawn.

-   The t-test requires the same assumption of random, independent samples, but it also assumes (1) the normality of the populations or (2) the samples are large enough.

-   The MWW test is good for nonsymmetric data.

-   The downside is that the statistic does not use all information in the data, only ranks and so it loses statistical power.

## The paired-sample t-test and CI

So far we have assumed that the we have two independent samples, one from each population. We will now consider the comparison of two samples that are not independent bu are **paired**. In a paired design, the observations $(Y_{1},Y_{2})$ occur in pairs; the observational units in a pair are linked.

Pairing in an experimental design can serve to reduce bias, to increase precision, or
both. Usually the primary purpose of pairing is to increase precision.

For example: consider testing whether a fertilizer increases plant growth using the same plants measured twice.

12 tomato plants are measured for height after 2 weeks under standard conditions. Then all 12 plants receive a new fertilizer for 2 more weeks and height is measured again. Because each plant is measured before and after, observations are paired.

$H_{0}: \mu_{D}=\mu_{1}-\mu_{2}=0$

$H_{A}: \mu_{D}=\mu_{1}-\mu_{2} \neq 0$.

```{r}
# Paired t-test example data (plant heights in cm)
dat <- data.frame(
  plant  = 1:6,
  before = c(12.1, 11.8, 13.0, 10.9, 12.5, 11.2),
  after  = c(14.0, 13.2, 14.1, 12.0, 13.7, 12.6)
)

dat

```

The test statistic is $$\frac{\bar{D}-0}{SE_{\bar{D}}}=\frac{\bar{D}-0}{SD_{D}/\sqrt{n}}$$ which is distributed as a t-distribution with $df=n-1$.

A $(1-\alpha)%$ confidence interval for $\mu_{D} is$$\bar{D}\pm t_{n-1,\alpha/2} SE_{\bar{D}}$\$ In our example this is:

```{r}
# CI by hand
d<-mean(dat$after-dat$before)
sd<-sd(dat$after-dat$before)
se<-sd/sqrt(6)
se
tval<-qt(.975,5)
d-tval*se
d+tval*se
print(paste("95% CI is: (",d-tval*se,d+tval*se,")"))


# Paired t-test
t.test(dat$after, dat$before, paired = TRUE)

```

What is the difference between a paired and an unpaired test?

```{r}
# CI by hand
d<-mean(dat$after-dat$before)
se1<-sd(dat$after)/sqrt(6)
se2<-sd(dat$before)/sqrt(6)
se_unpooled<-sqrt(se1^2+se2^2)
df<-((se1^2+se2^2)^2)/(((se1^2)^2 / 5)+((se2^2)^2 / 5))
tval<-qt(.975,df)
d-tval*se
d+tval*se
print(paste("95% CI is: (",d-tval*se_unpooled,d+tval*se_unpooled,")"))


# unpaired two-sided
t.test(dat$after, dat$before, paired = FALSE)

```

A paired experiment reduces the noise, i.e., variation in outcomes due to other variables.
