---
title: "Logistic Regression"
format: html
bibliography: references.bib
execute:
  echo: true
  warning: false
  message: false
  resources:
  - Lec20_annotated.pdf
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

<!-- Flex container aligns title and button with text margin -->

::: header-row
<h1 class="page-title">

</h1>

<a href="Lec20_annotated.pdf" class="download-btn" download> ⬇ Download Annotated Slides </a>
:::

Logistic regression is a special type of nonlinear regression developed for a binary response variable ($Y$), for example: 

- Disease status: yes/no  
- Treatment response: responder/non-responder  
- Smoking: current smoker/not  
- Mortality: dead/alive  

A common use for logistic regression is to fit a dose-response curve, where mortality (or survival) of individuals (the response) is plotted agins the concentration of a adrug, toxin, or other chemical.

If we tried linear regression on a 0/1 outcome, predicted values could fall outside $([0,1]$ and error assumptions would be violated. Logistic regression models the **probability** in a way that stays in $[0,1]$.

---

## The logistic model

Let $Y \in \{0,1\}$ be the outcome and $X$ be a predictor (or a set of predictors). Define:

$$
p(x) = P(Y=1 \mid X=x)
$$

Logistic regression uses the **logit link**:

$$
\text{logit}(p) = \log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x
$$

Equivalently, the predicted probability is:

$$
p(x) = \frac{1}{1 + \exp\left(-(\beta_0 + \beta_1 x)\right)}
$$

This creates an S-shaped curve mapping real numbers to probabilities.


Example:
```{r}
beta0 <- -6
beta1 <- 0.2
x <- seq(0, 60, by = 1)
p <- 1/(1 + exp(-(beta0 + beta1*x)))

plot(x, p, type = "l", lwd = 2,
     xlab = "x", ylab = "P(Y=1 | x)",
     main = "Logistic regression S-curve")
abline(h = c(0, 0.5, 1), lty = 3, col = "gray70")
```



### Log-odds interpretation

$$
\beta_1 = \text{change in log-odds of } Y=1 \text{ for a 1-unit increase in } x
$$

### Odds ratio interpretation

Exponentiating gives an **odds ratio (OR)**:

$$
\exp(\beta_1) = \text{multiplicative change in odds for a 1-unit increase in } x
$$

- If $\exp(\beta_1)=1$: no association  
- If $\exp(\beta_1)>1$: higher odds as $x$ increases  
- If $\exp(\beta_1)<1$: lower odds as $x$ increases  

**Example:** If $\exp(\beta_1)=1.50$, each 1-unit increase in $x$ is associated with **50% higher odds** of $Y=1$, holding other variables constant.

---

## Multiple logistic regression

With multiple predictors:

$$
\text{logit}(p) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k
$$

Interpretation of $\exp(\beta_j)$: adjusted odds ratio for $x_j$, **holding all other predictors fixed**.

This is often used to adjust for confounding in observational studies.


For example, we might model the probability of developing diabetes as a function of BMI and age
```{r}
#| warning: false
library("dplyr")
library(NHANES)
NHANES<-NHANES %>%
  mutate(has_diabetes=as.numeric(Diabetes == "Yes"))

par(mfrow=c(1,2))
    plot(NHANES$Age,NHANES$has_diabetes, xlab="Age",ylab="Diabetes status")
    plot(NHANES$BMI,NHANES$has_diabetes, xlab="BMI",ylab="Diabetes status")
    

```
We see that the probability that a subject has diabetes tends to increase as both a function of age and BMI.

Which variable is more important: Age or BMI? We can use a multiple logistic regression model to model the probability of diabetes as a function of both predictors

```{r}
logreg<-glm(has_diabetes~BMI+Age, family="binomial",data=NHANES)
summary(logreg)
```

Both predictors seem to be important. To interpret the findings, we might consider a visual display of predicted probabilities as 

```{r}
library(modelr)
fake_grid<- data_grid(
  NHANES,
  Age =seq_range(Age,100),
  BMI=seq_range(BMI,100)
)

y_hats<- fake_grid %>%
  mutate(y_hat=predict(logreg,newdata=.,type="response"))
head(y_hats,1)


```

## Interpreting coefficients

The predicted probability from the model is given by:
$$
\pi_{i}=P(y_{i}=1\mid Age_{i},BMI_{i})=\frac{e^{\beta_{0}+\beta_{1}Age_{i}+\beta_{2}BMI_{i}}}{1+e^{\beta_{0}+\beta_{1}Age_{i}+\beta_{2}BMI_{i}}}
$$
Let's consider a hypothetical 0 year old with a BMI of 12.0. Their predicted probability of having diabetes would be calculated as a function of the regression coefficients:

```{r}
linear_component<-c(1,12.9,0)%*%coef(logreg)
exp(linear_component)/(1+exp(linear_component))

```
The predicted probability is very small, what about a 60 year old with a BMI of 25?


```{r}
linear_component<-c(1,25,60)%*%coef(logreg)
exp(linear_component)/(1+exp(linear_component))

```

The predicted probability is now 9.2%

---

## Categorical predictors

Suppose sex is coded as:

- Female = 0 (reference)
- Male = 1

Then:

$$
\exp(\beta_{\text{male}}) = \frac{\text{odds of } Y=1 \text{ for males}}{\text{odds of } Y=1 \text{ for females}}
$$

For predictors with more than two categories, use indicator (dummy) variables with a reference group.



