---
title: "Probability"
format: html
bibliography: references.bib
---

The concept of probability is important in almost every field of science, including biology. Probability is the foundation of statistical inference. The concept of probability rests on the notion of a **random experiment** that has two or more possible outcomes whose occurrence annot be predicted with certainty.


For example, consider the experiment of rolling a pair of dice. The set of possible outcomes (sample space) is (1,1),(1,2),...,(6.6). An **event** is a subset of the sample space. 

Example 1: The event is $E_{1}=(1,1)$.


Example 2: The event is $E_{2}=$"The sum of the two rolls is 5"



The **probability** of an event is the proportion of all random experiments in which the specified event occurs when the same experiment is repeated independently over and over again. In the long run, that is, if an infinite number of experiments is carried out, the probability of an event is the fraction of experiments in which the even occurs. Probability ranges between 0 an 1.


A useful tool for thinking about probabilities of events is with a Venn diagram. The area of the diagram represents all possible outcomes of an experiment, and we can represent various events as areas within the diagram.

**Activity 4**

Two events are **mutually exclusive** if both cannot occur simultaneously. For example $E_{1}$ and $E_{2}$ defined above are mutually exclusive and
$$ P(E_{1} \text{ and } E_{2})=0$$

##Probability distribution

A simple way to think about a **probability distribution** is as a compact description of many numbers. For example, imagine that all the heights of men in a population have been measured, and you want to describe or explain those measures to someone else. Suppose all these measures are contained in the following dataset:

```{r}
data(father.son, package="UsingR")
x <- father.son$fheight
```

One way to describe this information is to list all elements in the dataset, or to show a random sample of 10 elements from the list of 1,078 total measures.

```{r}
length(x)
sample(x,10)
```

Another, more compact and descriptive way to represent the data is by using a **histogram**. Histograms show the proportion of values within intervals, $(a, b)$. In the previous example, plotting the heights as bars is more illustrative because we are usually more interested in intervals, such as the percentage of people whose heights fall between 70 and 71 inches. Histograms are also useful for describing different families of distributions. Here is the histogram of the height data:

```{r}
hist(x, xlab="Height (in inches)", main="Adult men heights",freq=TRUE)
```

With this plot, we can describe the number of observations within any given interval. For example, there are over 70 individuals taller than 6 feet (72 inches).

Summarizing a list of numbers is a powerful way to describe the distribution or, more importantly, the possible outcomes of a random experiment. We may not know all possible outcomes, so instead of describing proportions, we describe probabilities. For instance, the probability of picking a random height within the interval $(a,b)$ is denoted with $Pr(a \leq X \leq b)$.

In summary, a **probability distribution** describes the probabilities of each of the possible outcomes of a random experiment. Some probability distributions can be described mathematically, while others are just a list of possible outcomes and their probabilities (as in the heights example).

The **cumulative distribution function (CDF)**, represented by the following notation: $F(a) \equiv \Pr(X \leq a)$ \[and the proportion of numbers bigger than $a$ is computed as: $\Pr(X \gt a)=1-\Pr(X \leq a)=1-F(a)$\].

** This or that, the addition rule



** Independence and multiplication rule



** Dependent events



** Conditional probability and Bayes' theorem
 

## Discrete probability models

In many contexts-such as molecular biology-we count events (e.g., codons, DNA reads, CG diagrams), producing discrete variables, unlike continuous measures such as mass or intensity. Some variables are binary (e.g., mutation: yes/no), while others have multiple levels (e.g., genotypes AA, Aa, aa) or even many categories (e.g., bacterial types or 64 codons).

To summarize categorical data, we tally the frequency of each level. In R, these are stored as factors, which make tabulation straightforward, as shown previously. For example, here we capture the different blood genotypes for 19 subjects in a vector which we tabulate.

```{r}
genotype = c("AA","AO","BB","AO","OO","AO","AA","BO","BO",
             "AO","BB","AO","BO","AB","OO","AB","BB","AO","AO")
table(genotype)
```

If the order in which the data are observed does not matter, we call the random variable exchangeable. In this case, the vector of frequencies is sufficient to capture all the relevant information in the data, providing an efficient way to summarize it.

### Bernoulli trials

The **Bernoulli distribution** is the simplest probability model, based on a so-called Bernoulli random variable, which can take only two outcomes: failure or success. The key idea is that you have a parameter $p$, which represents the probability of success, and the complementary event is called probability of failure $1-p$:

***Probability distribution:***

$P(X=1) = p$\
$P(X=0) = 1-p$

For a fair coin, p = 0.5. For a biased coin that lands heads 70% of the time, p = 0.7.

::: {.callout-note icon="false"}
**Example:** Coin tosses\
Suppose we want to simulate a sequence of 15 fair tosses using R. To get the outcome of 15 tosses of a fair coin (p = 0.50), we write:
:::

```{r}
rbinom(15, prob = 0.5, size = 1)
```

Now observe the outcome when using a biased coin that lands heads 70% of the time:

```{r}
rbinom(15, prob = 0.7, size = 1)
```

*Question.* Note that when you call the function above several times, you get different outcomes. Why is that?

***Properties***

The **mean**, or **expected value**, of a Bernoulli distribution is given by $p$, and the **variance** is defined as $p(1-p)$, so the variance is maximized when p = 0.5, and it goes to 0 as p approaches 0 or 1. The Bernoulli distribution is very important because it represents the building block of many other distributions.

### Binomial success counts

In the previous example, if we are only interested in the number of heads ($X=1$), we just take the sum of the cells in the output vector, then reporting a single number instead of a binary vector. In R, we can do this as follows:

```{r}
rbinom(1, prob = 0.7, size = 15)
```

The sum of multiple independent (or exchangeable) Bernoulli trials is called the **Binomial distribution** (in the example above, how many heads in 15 flips). That is, the number of successes in 15 Bernoulli trials with a probability of success of 0.7 is called a **binomial** random variable, or a random variable that follows the B(15, 0.7) distribution. We can generate binomial samples in R as follows:

```{r}
set.seed(384569724)
rbinom(1, prob = 0.7, size = 15)
```

*Question.* Repeat this function call 10 times. What seems to be the most common outcome?

*Solution.* The most frequent value is 10.5$\approx 11$. In fact, the theoretical proportion of times that we expect 10.5 to appear is the value of the probability that $X=4$ if $X$ follows B(15,0.7).

The complete **probability mass distribution** is available by typing:

```{r}
probabilities = dbinom(0:15, prob = 0.7, size = 15)
round(probabilities, 2)
```

A barplot for this distribution in R, is as follows:

```{r}
barplot(probabilities, names.arg = 0:15, col = "black", main = "Theoretical distribution of B(15, 0.7).")
```

Mathematically, the theoretical distribution of the binomial distribution is

***Probability distribution:***

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}, \quad k = 0, 1, \ldots, n,$$ where $n$ is the number of trials, $k$ is the number of successes, $p$ is the probability of success, and $\binom{n}{k} = \frac{n!}{k!(n-k)!}$ is the binomial coefficient.

***Properties***

The **mean** of a Binomial distribution is given by $n \cdot p$ (in the figure, the highest bar around $x \approx 11$, or $n(=15)\cdot p(=0.7)$, and the **variance** is defined as $np(1-p) = (15)(0.7)(0.3) = 3.15$.

### Poisson distributions

When the probability of success $p$ is small and the number of trials $n$ is large, the Binomial distribution can be approximated by a **Poisson distribution** with parameter $\lambda = np$, which provides a simple model for describing the data.

***Probability distribution:***

$$P(X = k) = \frac{e^{-\lambda} \lambda^k}{k!}, \quad k = 0, 1, 2, \ldots,$$ where $k$ is the number of events, $\lambda$ is the average number of events per interval, and $e\approx 2.718$ is the Euler's number.

***Properties***

An interesting characteristic of the Poisson distribution is that the **mean** and the **variance** are both equal to $\lambda = np$.

::: {.callout-note icon="false"}
**Example:** Human Immunodeficiency Virus (HIV)\
Consider a probability model for HIV mutations, where the data generation process is described by a Poisson model. That is, mutations occur at a constant rate of $p=5 x 10^{-4}$ per nucleotide per replication cycle, independently across positions. Given a genome size of approximately \$n\$ nucleotides, the total number of mutations per cycle follows a $\text{Poisson}(\lambda)$ distribution.

As the genome size of HIV is about $n=10^4$, this model predicts an expected number of mutations close to 5 $(np=10^4 x 5 x 10^{-4})$ per replication cycle, with standard error $\sqrt{\lambda}$. For example, to find the probability of exactly 3 mutations under a $\text{Poisson}(5)$ model, we can use R:

```{r}
dpois(x=3, lambda=5)
```

This means that the chance of seeing exactly three events is about 0.14.

Now we can visualize the distribution of the values from a particular range, from 0 to 12 for example, we proceed as follows:

```{r}
y <- dpois(x=0:12, lambda=5)
barplot(y, names.arg = 0:12, col = "#FFCCAA",  main = "Probabilities of seeing 0 to 12 mutations, modeled as a Poisson(5)")
```

This example shows that the Poisson distribution is a good model for rare events such as mutation (observe that $p$ is very low).
:::

### Multinomial distributions

In the Binomial model, we had only two possible outcomes, success and failure, with probabilities $p$ and $1-p$, respectively. How do we extend that model for more than 2 outcomes? For example, when studying counts of the four nucleotides in DNA: $\{A, C, G, T\}$. We use a **Multinomial model**, where we assign different probabilities to the different outcomes. In the case of the four DNA nucleotides, for example, we can represent the respective probabilities as $p_{A}$, $p_{C}$, $p_{G}$, and $p_{T}$, where the probabilities sum to 1: $p_{A}+p_{C}+p_{G}+p_{T}=1$, as in the Binomial model.

The mathematical formulation for the Multinomial model for $m$ counts $(x_{1}, x_{2}, \ldots, x_{m})$ for outcomes of $n$ draws with probabilities $p_{1}, p_{2}, \ldots, p_{m}$ is as follows:

***Probability distribution:***

$$\begin{align}
P(x_{1}, x_{2}, \ldots, x_{m}) &= \frac{n!}{\prod_{i=1}^{m} x_{i}!} \prod_{i=1}^{m} p_{i}^{x_{i}} \\
&= \binom{n}{x_{1}, x_{2}, \ldots, x_{m}} p_{1}^{x_{1}} p_{2}^{x_{2}} \cdots p_{m}^{x_{m}}
\end{align}$$

To generate draws that are consistent with this model, we use R commands to generate vector counts with each having equal probability of selection (1/4 in the example of 4 nucleotides). For example, suppose we have 8 characters of four different equally likely types:

```{r}
pvec = rep(1/4, 4)
t(rmultinom(1, prob = pvec, size = 8))
```

*Question* Try the code without using the t() function; what does t stand for?

Reference: @holmes2018modern
