---
title: "Central limit theorem"
---

![Statistics](/Figures/pop_cats){fig-align="center" width="95%"}

In the previous lectures, the knowledge of both the generative model and the values of the **parameters** provided us with probabilities we could use for decision making – for instance, what is the more likely height of somebody chosen at random. In many real situations, neither the generative model nor the parameters are known, and we will need to estimate them using the data we have collected. 


Statistical modeling works from the data upwards to a model that might plausibly explain the data. This upward-reasoning step is called statistical inference. 
![Statistical Inference](/Figures/inference){fig-align="center" width="95%"}

Let's first distinguish between population parameters and statistics (functions of our observed data) that have similar names:

| Parameter | Statistic |
|-----------------------------------------------|-------------------------|
| Expected value: $E(X)=\sum^{N}_{i}x_{i}P(X=x_{i})$ | Sample mean: $\bar{x}=\frac{\sum^{n}_{i}x_{i}}{n}$ |
| Variance: $Var(X)=\sum^{N}_{i=1}(x_{i}-E(X))^2P(X=x_{i})$ | Sample variance: $\sqrt{\frac{\sum^{n}_{i=1}(x_{i}-\bar{x})^2}{n-1}}$ |
| Probability: $P(X=a)$ | Empirical proportion: $\frac{\#a}{n}$|


The law of large numbers states that the sample mean will converge to the expected value as $n \rightarrow \infty$. For example:

```{r}
x<-rnorm(10,0,1)
mean(x)
x<-rnorm(100,0,1)
mean(x)
x<-rnorm(1000,0,1)
mean(x)
x<-rnorm(10000,0,1)
mean(x)
```
Clearly, different samples will give us different sample means. How much do we expect them to vary? We can compute the sample standard deviation of the sample mean as follows:

```{r}
vector_means<-rep(0,1000)
for (j in 1:1000){
  vector_means[j]<-mean(rnorm(10,0,1))
}
sqrt(var(vector_means))
hist(vector_means)

```
The theoretical standard deviation of the sample mean (called **standard error**) is
$$SD(\bar{X})=\frac{SD_{X}}{\sqrt{n}}$$
Why?

```{r}
1/sqrt(10)
```

Notice that the **empirical** distribution of the mean remarkably looks like the Gaussian distribution. Let's try now to analyze the distribution of the mean of a binomial distribution.

```{r}
vector_means<-rep(0,1000)
for (j in 1:1000){
  vector_means[j]<-mean(rbinom(10,size=5,0.5))
}
sqrt(var(vector_means))
hist(vector_means)
sqrt(5*.5*.5)/sqrt(10)

```
Remarkably this histogram also looks Gaussian!

## Central limit theorem
Let $X_{1},\ldots,X_{n}$ be a random sample (independent and identically distributed) from a population with mean $E(X)$ and standard deviation $SD_{X}$, then
$$Z=\frac{\bar{X}-E(X)}{SD_{x}/\sqrt{n}} $$ and $Z \approx N(0,1)$.

Why do we care?


