# Stats 141 — Review Problems


1. In a study of stress levels in U.S. Army recruits stationed in Iraq, researchers obtained a complete list of the names of recruits in Iraq at the time of the study. They listed the recruits alphabetically and then numbered them consecutively. One hundred  random numbers between one and the total number of recruits were then generated using a random-number generator on a computer (without replacement). The 100 recruits whose numbers corresponded to those generated by the computer were interviewed for the study.

(a) What is the population of interest in this study?

(b) Is the sample affected by sampling error?

(c) What are the main advantages of random sampling in this example?

(d) What effect would a larger sample size have had on sampling error?


2. A biologist measures the body mass (in grams) of a sample of 10 adult mice:

$$22, 25, 19, 30, 24, 28, 21, 26, 23, 27$$

(a) Compute the sample mean, median, and standard deviation.

(b) Is this distribution symmetric, right-skewed, or left-skewed? Justify your answer using the mean and median.

(c) One mouse has a mass of 30 g. Is this an outlier? Use the IQR rule (outlier if beyond $Q1 - 1.5 \times IQR$ or $Q3 + 1.5 \times IQR$).


3. A study was carried out in women to determine whether the psychological consequences of having an abortion differ from those experienced by women who have lost their fetuses by other causes at the same stage of pregnancy.
(a) Which is the explanatory variable in this study, and which is the response variable?

(b) Was this an observational study or an experimental study? Explain.

(c) What is the main threat to validity of this study? Name a specific confounder.

(d) You find a correlation between coffee consumption and miscarriage. Can you conclude that drinking coffee **causes** miscarriage? 

(e) What type of plot would you use to show whether there is correlation between two continuous variables?


4. A genetic test for a rare disease has the following properties:

- Prevalence of the disease: $P(D) = 0.01$
- Sensitivity (true positive rate): $P(+ \mid D) = 0.95$
- Specificity (true negative rate): $P(- \mid D^c) = 0.90$

(a) What is the probability of testing positive, $P(+)$?

(b) Use Bayes' theorem to compute the **positive predictive value** $P(D \mid +)$.

(c) Interpret the result in plain language. Is a positive test strong evidence of disease?



5. Researchers measure blood glucose (mg/dL) in a random sample of $n = 36$ adult patients. They find:

- Sample mean: $\bar{x} = 102$ mg/dL
- Sample standard deviation: $s = 18$ mg/dL

(a) Compute a 95% confidence interval for the population mean blood glucose

(b) What does this confidence interval mean? Choose the correct interpretation:

   i. There is a 95% probability that the true mean lies in this interval.
   ii. If we repeated this study many times, 95% of such intervals would contain the true mean.
   iii. 95% of individual patients have blood glucose between these values.

(c) How would the interval change if $n = 144$ instead of 36?



6. A pharmaceutical company claims their drug reduces systolic blood pressure by more than 10 mmHg. In a clinical trial of $n = 25$ patients, the mean reduction was $\bar{x} = 12.4$ mmHg with $s = 6$ mmHg.

(a) State the null and alternative hypotheses.

(b) Compute the test statistic and p-value.

(c) At $\alpha = 0.05$, what do you conclude?

(d) What is a Type I error in this context?



7. Researchers compare the fasting insulin levels (μU/mL) in two groups:

- **Group A** (low-carb diet): $n_A = 20$, $\bar{x}_A = 8.2$, $s_A = 2.4$
- **Group B** (standard diet): $n_B = 20$, $\bar{x}_B = 10.5$, $s_B = 3.1$

(a) Conduct a two-sample $t$-test (assume unequal variances) at $\alpha = 0.05$.

(b) A collaborator suggests using a Wilcoxon-Mann-Whitney test instead. When would this be preferable?



8. A researcher screens 500 genes for differential expression between cancer and healthy tissue. Using individual t-tests at $\alpha = 0.05$, they find 40 genes with $p < 0.05$.

(a) How many false positives would you expect if none of the genes are truly differentially expressed?

(b) The researcher applies the **Bonferroni correction**. What is the adjusted significance threshold?

(c) They instead use the **Benjamini-Hochberg (BH) procedure** with FDR = 0.05. Explain conceptually what this controls.



9. In a study of 200 patients, researchers record disease outcome (improved / not improved) and treatment group (drug / placebo):

|               | Improved | Not Improved | Total |
|---------------|----------|--------------|-------|
| Drug          | 70       | 30           | 100   |
| Placebo       | 50       | 50           | 100   |
| **Total**     | 120      | 80           | 200   |

(a) Compute the chi-square test statistic and p-value.

(b) At $\alpha = 0.05$, is there a significant association between treatment and outcome?




10. The following data show the relationship between daily average temperature (°C) and the growth rate (mm/day) of a plant species. We fit a simple linear regression of growth on temperature. 

```{r}
temp   <- c(10, 15, 18, 20, 22, 25, 28, 30)
growth <- c(1.2, 1.8, 2.3, 2.7, 3.0, 3.5, 3.9, 4.1)
model <- lm(growth ~ temp)
summary(model)
```

(a) State the (fitted) prediction model and your assumptions.

(b) Interpret the coefficients, are they significant?

(c) What is the predicted growth rate at 24°C?

(d) Report $R^2$ and interpret it.


11. A study models the concentration of a protein biomarker (ng/mL) as a function of patient age (years) and BMI.

```{r}
set.seed(1)
n   <- 100
age <- round(runif(n, 30, 70))
bmi <- round(rnorm(n, 27, 4), 1)
biomarker <- 2.0 + 0.05 * age + 0.10 * bmi + rnorm(n, 0, 1.2)

fit_simple   <- lm(biomarker ~ age)
fit_multiple <- lm(biomarker ~ age + bmi)
```

(a) In the simple regression of biomarker on age alone, what is the estimated slope for age?

(b) In the multiple regression including BMI, does the slope for age change? Why?

(c) What is the interpretation of the BMI coefficient in the multiple regression model?



12. Using the NHANES dataset, a logistic regression models the probability of diabetes as a function of BMI and Age:

$$\text{logit}(p) = -8.08 + 0.094 \times \text{BMI} + 0.057 \times \text{Age}$$

(a) Compute the predicted probability of diabetes for a 55-year-old with BMI = 32.

(b) Compute and interpret the odds ratio for BMI.

(c) Is BMI or Age a stronger predictor in this model on the log-odds scale? How would you compare them more fairly?

(d) A patient's BMI increases from 28 to 38 (a 10-unit increase). By what factor do their odds of diabetes change?



13. A clinical trial follows 8 patients after treatment. Their survival times (months) and event status (1 = died, 0 = censored) are:

```{r}
times  <- c(3, 5, 6, 6, 8, 10, 12, 15)
events <- c(1, 1, 1, 0, 1, 0,  1,  0)

library(survival)
fit <- survfit(Surv(times, events) ~ 1)
plot(fit,
     xlab = "Time (months)",
     ylab = "Survival probability",
     main = "Kaplan-Meier Curve",
     lwd  = 2, col = "steelblue",
     conf.int = TRUE)
```



(a) What is the estimated probability of surviving beyond 10 months?

(b) One patient was censored at month 6. What does this mean, and how does the KM estimator handle it?


Formulas:
unpooled SE: $$
\mathrm{SE}\!\left(\bar X_1-\bar X_2\right)
= \sqrt{\frac{s_1^{2}}{n_1}+\frac{s_2^{2}}{n_2}}
$$


degrees of freedom: $$ 
\mathrm{df}
=
\frac{\left(\frac{s_T^2}{n_T}+\frac{s_C^2}{n_C}\right)^2}
{\frac{\left(\frac{s_T^2}{n_T}\right)^2}{n_T-1}
+
\frac{\left(\frac{s_C^2}{n_C}\right)^2}{n_C-1}}.
$$

Goodness of fit test

$$
\chi^{2}=\sum_{i}\frac{(\text{Observed}_{i}-\text{Expected}_{i})^2}{\text{Expected}_{i}}
$$