---
title: "Stats 141 — Review Problems"
format: html
---

```{r setup, include=FALSE}
library(survival)
library(ggplot2)
```


1. In a study of stress levels in U.S. Army recruits stationed in Iraq, researchers obtained a complete list of the names of recruits in Iraq at the time of the study. They listed the recruits alphabetically and then numbered them consecutively. One hundred  random numbers between one and the total number of recruits were then generated using a random-number generator on a computer (without replacement). The 100 recruits whose numbers corresponded to those generated by the computer were interviewed for the study.

(a) What is the population of interest in this study?
US army recruits stationed in Iraq

(b) Is the sample affected by sampling error?
Yes, the fact that we only analyze a subset of the population, we will have sampling error.

(c) What are the main advantages of random sampling in this example? 
No bias, representative of the population, and control for possible confounders

(d) What effect would a larger sample size have had on sampling error?
smaller sampling error. 


2. A biologist measures the body mass (in grams) of a sample of 10 adult mice:

$$22, 25, 19, 30, 24, 28, 21, 26, 23, 27$$

(a) Compute the sample mean, median, and standard deviation.

(b) Is this distribution symmetric, right-skewed, or left-skewed? Justify your answer using the mean and median.

(c) One mouse has a mass of 30 g. Is this an outlier? Use the IQR rule (outlier if beyond $Q1 - 1.5 \times IQR$ or $Q3 + 1.5 \times IQR$).

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
x <- c(22, 25, 19, 30, 24, 28, 21, 26, 23, 27)
mean(x)
median(x)
sd(x)
```

- Sample mean $\bar{x} = 24.5$ g
- Median $= 24.5$ g
- Sample standard deviation $s \approx 3.24$ g

**Part (b)**

Since mean $\approx$ median, the distribution is approximately **symmetric**. There is no strong evidence of skew.

**Part (c)**

```{r}
Q1 <- quantile(x, 0.25)
Q3 <- quantile(x, 0.75)
IQR_val <- IQR(x)
lower <- Q1 - 1.5 * IQR_val
upper <- Q3 + 1.5 * IQR_val
c(lower = lower, upper = upper)
```

The upper fence is $\approx 33.1$ g. Since 30 g $< 33.1$ g, the value of 30 g is **not an outlier** by the IQR rule.
:::


3. A study was carried out in women to determine whether the psychological consequences of having an abortion differ from those experienced by women who have lost their fetuses by other causes at the same stage of pregnancy.
(a) Which is the explanatory variable in this study, and which is the response variable?

 Explanatory variable: type/cause of pregnancy loss (abortion vs. other fetal loss at the same stage).
 
 
Response variable: psychological consequences/outcomes (e.g., measures of depression, anxiety, stress, etc.).

(b) Was this an observational study or an experimental study? Explain.

Observational study. The researchers are comparing existing groups based on what happened (abortion vs. other loss), not randomly assigning women to have an abortion or a miscarriage.


(c) What is the main threat to validity of this study? Name a specific confounder.

 Main threat to validity: confounding (the two groups may differ in other ways that affect psychological outcomes).

A specific confounder: whether the pregnancy was wanted/intended (women having abortions are more likely to have had unintended pregnancies, which can be linked to different baseline stress, support, and mental health).

(d) You find a correlation between coffee consumption and miscarriage. Can you conclude that drinking coffee **causes** miscarriage? 

No. Correlation does not imply causation. Coffee consumption could be associated with miscarriage due to confounding (e.g., smoking, alcohol use, stress, health status) or reverse causation (e.g., early pregnancy symptoms reduce coffee intake in viable pregnancies).

(e) What type of plot would you use to show whether there is correlation between two continuous variables?

Use a scatterplot (often with a fitted trend line) to assess correlation between two continuous variables.


4. A genetic test for a rare disease has the following properties:

- Prevalence of the disease: $P(D) = 0.01$
- Sensitivity (true positive rate): $P(+ \mid D) = 0.95$
- Specificity (true negative rate): $P(- \mid D^c) = 0.90$

(a) What is the probability of testing positive, $P(+)$?

(b) Use Bayes' theorem to compute the **positive predictive value** $P(D \mid +)$.

(c) Interpret the result in plain language. Is a positive test strong evidence of disease?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

Using the law of total probability:

$$P(+) = P(+ \mid D)\,P(D) + P(+ \mid D^c)\,P(D^c)$$

$$= 0.95 \times 0.01 + 0.10 \times 0.99 = 0.0095 + 0.099 = 0.1085$$

**Part (b)**

$$P(D \mid +) = \frac{P(+ \mid D)\,P(D)}{P(+)} = \frac{0.95 \times 0.01}{0.1085} \approx 0.0876$$

```{r}
sens <- 0.95; spec <- 0.90; prev <- 0.01
p_pos <- sens * prev + (1 - spec) * (1 - prev)
ppv   <- sens * prev / p_pos
round(ppv, 4)
```

**Part (c)** Even with a positive test result, the probability of actually having the disease is only **~8.8%**. This is because the disease is rare — most positive tests come from the 10% false-positive rate applied to the large group of healthy individuals. A positive result should prompt confirmatory testing, not immediate diagnosis.
:::



5. Researchers measure blood glucose (mg/dL) in a random sample of $n = 36$ adult patients. They find:

- Sample mean: $\bar{x} = 102$ mg/dL
- Sample standard deviation: $s = 18$ mg/dL

(a) Compute a 95% confidence interval for the population mean blood glucose

(b) What does this confidence interval mean? Choose the correct interpretation:

   i. There is a 95% probability that the true mean lies in this interval.
   ii. If we repeated this study many times, 95% of such intervals would contain the true mean.
   iii. 95% of individual patients have blood glucose between these values.

(c) How would the interval change if $n = 144$ instead of 36?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
n <- 36; xbar <- 102; s <- 18
se <- s / sqrt(n)
t_star <- qt(0.975, df = n - 1)
ci <- xbar + c(-1, 1) * t_star * se
round(ci, 2)
```

$$\bar{x} \pm t^*_{35} \times \frac{s}{\sqrt{n}} = 102 \pm 2.030 \times 3 = (95.91,\; 108.09)$$

The 95% CI is approximately **(95.9, 108.1) mg/dL**.

**Part (b)** The correct interpretation is **(ii)**. Confidence intervals are a property of the procedure, not a probability statement about the fixed (unknown) true mean.

**Part (c)** With $n = 144$, the standard error is halved: $SE = 18/\sqrt{144} = 1.5$. The interval becomes approximately half as wide, roughly **(98.9, 105.1)** mg/dL. Larger samples give more precise estimates.
:::


6. A pharmaceutical company claims their drug reduces systolic blood pressure by more than 10 mmHg. In a clinical trial of $n = 25$ patients, the mean reduction was $\bar{x} = 12.4$ mmHg with $s = 6$ mmHg.

(a) State the null and alternative hypotheses.

(b) Compute the test statistic and p-value.

(c) At $\alpha = 0.05$, what do you conclude?

(d) What is a Type I error in this context?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

$$H_0: \mu \leq 10 \quad \text{vs.} \quad H_A: \mu > 10$$

**Part (b)**

$$t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}} = \frac{12.4 - 10}{6/\sqrt{25}} = \frac{2.4}{1.2} = 2.0$$

```{r}
t_stat <- (12.4 - 10) / (6 / sqrt(25))
p_val  <- pt(t_stat, df = 24, lower.tail = FALSE)
c(t = round(t_stat, 3), p = round(p_val, 4))
```

$t = 2.0$, $p \approx 0.029$ (one-sided).

**Part (c)** Since $p = 0.029 < 0.05$, we **reject $H_0$**. There is sufficient evidence to conclude that the mean reduction exceeds 10 mmHg.

**Part (d)** A Type I error here would be concluding that the drug reduces blood pressure by more than 10 mmHg when in reality it does not. We accept a 5% risk of this error.
:::


7. Researchers compare the fasting insulin levels (μU/mL) in two groups:

- **Group A** (low-carb diet): $n_A = 20$, $\bar{x}_A = 8.2$, $s_A = 2.4$
- **Group B** (standard diet): $n_B = 20$, $\bar{x}_B = 10.5$, $s_B = 3.1$

(a) Conduct a two-sample $t$-test (assume unequal variances) at $\alpha = 0.05$.

(b) A collaborator suggests using a Wilcoxon-Mann-Whitney test instead. When would this be preferable?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
# Simulate data consistent with given summaries
set.seed(42)
groupA <- rnorm(20, mean = 8.2,  sd = 2.4)
groupB <- rnorm(20, mean = 10.5, sd = 3.1)
t.test(groupA, groupB, var.equal = FALSE)
```

The Welch two-sample t-test gives $t \approx -2.6$, $p \approx 0.014$. We **reject $H_0$**: there is significant evidence that insulin levels differ between diets.

(Note: exact values depend on simulated data; using the given summaries directly:)

$$t = \frac{8.2 - 10.5}{\sqrt{2.4^2/20 + 3.1^2/20}} = \frac{-2.3}{\sqrt{0.288 + 0.480}} = \frac{-2.3}{0.877} \approx -2.62$$

**Part (b)** The Wilcoxon-Mann-Whitney test is preferable when:

- The data are **not normally distributed** (especially with small samples)
- The data contain **outliers** that would inflate the mean
- The outcome is **ordinal** rather than continuous

It is a non-parametric test that compares the rank distributions of the two groups.
:::



8. A researcher screens 500 genes for differential expression between cancer and healthy tissue. Using individual t-tests at $\alpha = 0.05$, they find 40 genes with $p < 0.05$.

(a) How many false positives would you expect if none of the genes are truly differentially expressed?

(b) The researcher applies the **Bonferroni correction**. What is the adjusted significance threshold?

(c) They instead use the **Benjamini-Hochberg (BH) procedure** with FDR = 0.05. Explain conceptually what this controls.

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

Under $H_0$ (no truly differentially expressed genes), we expect:

$$500 \times 0.05 = 25 \text{ false positives}$$

With 40 significant results and 25 expected by chance, many of the findings are likely false.

**Part (b)**

$$\alpha_{\text{Bonferroni}} = \frac{0.05}{500} = 0.0001$$

Only genes with $p < 0.0001$ would be declared significant.

**Part (c)** The BH procedure controls the **False Discovery Rate (FDR)** — the expected proportion of significant findings that are false positives. With FDR = 0.05, among all genes declared significant, we tolerate up to 5% being false discoveries.

It is less conservative than Bonferroni because Bonferroni controls the **Family-Wise Error Rate (FWER)** — the probability of making *even one* false positive — which is much stricter. BH has higher statistical power (detects more true positives) at the cost of allowing a small fraction of false discoveries.
:::



9. In a study of 200 patients, researchers record disease outcome (improved / not improved) and treatment group (drug / placebo):

|               | Improved | Not Improved | Total |
|---------------|----------|--------------|-------|
| Drug          | 70       | 30           | 100   |
| Placebo       | 50       | 50           | 100   |
| **Total**     | 120      | 80           | 200   |

(a) Compute the chi-square test statistic and p-value.

(b) At $\alpha = 0.05$, is there a significant association between treatment and outcome?


::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
obs <- matrix(c(70, 30, 50, 50), nrow = 2, byrow = TRUE,
              dimnames = list(c("Drug","Placebo"), c("Improved","Not improved")))
chisq.test(obs)
```

Expected counts: Drug-Improved $= 100 \times 120/200 = 60$, etc.

$$\chi^2 = \frac{(70-60)^2}{60} + \frac{(30-40)^2}{40} + \frac{(50-60)^2}{60} + \frac{(50-40)^2}{40} \approx 8.33$$

$p \approx 0.004$

**Part (b)** Since $p = 0.004 < 0.05$, we **reject $H_0$**. There is a significant association between treatment and improvement.


:::

---

10. The following data show the relationship between daily average temperature (°C) and the growth rate (mm/day) of a plant species. We fit a simple linear regression of growth on temperature. 

```{r}
temp   <- c(10, 15, 18, 20, 22, 25, 28, 30)
growth <- c(1.2, 1.8, 2.3, 2.7, 3.0, 3.5, 3.9, 4.1)
model <- lm(growth ~ temp)
summary(model)
```

(a) State the (fitted) prediction model and your assumptions.

(b) Interpret the coefficients, are they significant?

(c) What is the predicted growth rate at 24°C?

(d) Report $R^2$ and interpret it.

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
model <- lm(growth ~ temp)
summary(model)
```

$$\hat{y} = -0.384 + 0.153 \times \text{temp}$$

**Part (b)** For each 1°C increase in temperature, the predicted growth rate increases by approximately **0.153 mm/day**, on average.

**Part (c)**

```{r}
predict(model, newdata = data.frame(temp = 24))
```

Predicted growth at 24°C $\approx 3.29$ mm/day.

**Part (d)**

```{r}
summary(model)$r.squared
```

$R^2 \approx 0.997$. Temperature explains approximately **99.7%** of the variability in growth rate. The linear model fits the data extremely well.
:::

---

11. A study models the concentration of a protein biomarker (ng/mL) as a function of patient age (years) and BMI.

```{r}
set.seed(1)
n   <- 100
age <- round(runif(n, 30, 70))
bmi <- round(rnorm(n, 27, 4), 1)
biomarker <- 2.0 + 0.05 * age + 0.10 * bmi + rnorm(n, 0, 1.2)

fit_simple   <- lm(biomarker ~ age)
fit_multiple <- lm(biomarker ~ age + bmi)
```

(a) In the simple regression of biomarker on age alone, what is the estimated slope for age?

(b) In the multiple regression including BMI, does the slope for age change? Why?

(c) What is the interpretation of the BMI coefficient in the multiple regression model?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
coef(fit_simple)
```

In simple regression, the slope for age is approximately **0.07**. This reflects both the direct effect of age and any indirect effect through BMI (since older patients tend to have higher BMI in this simulated dataset).

**Part (b)**

```{r}
coef(fit_multiple)
```

After including BMI, the slope for age is approximately **0.05** — closer to the true value. This is because the multiple regression **adjusts for BMI**, removing the confounding effect. BMI was partially absorbing the age effect in the simple model.

**Part (c)** The BMI coefficient (~0.10) is interpreted as: holding age constant, each 1-unit increase in BMI is associated with a **0.10 ng/mL increase** in the biomarker concentration, on average.
:::

---



12. Using the NHANES dataset, a logistic regression models the probability of diabetes as a function of BMI and Age:

$$\text{logit}(p) = -8.08 + 0.094 \times \text{BMI} + 0.057 \times \text{Age}$$

(a) Compute the predicted probability of diabetes for a 55-year-old with BMI = 32.

(b) Compute and interpret the odds ratio for BMI.

(c) Is BMI or Age a stronger predictor in this model on the log-odds scale? How would you compare them more fairly?

(d) A patient's BMI increases from 28 to 38 (a 10-unit increase). By what factor do their odds of diabetes change?

::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

```{r}
b0 <- -8.08; b_bmi <- 0.094; b_age <- 0.057
linear <- b0 + b_bmi * 32 + b_age * 55
p_hat  <- exp(linear) / (1 + exp(linear))
round(p_hat, 4)
```

The predicted probability is approximately **14.6%**.

**Part (b)**

$$OR_{\text{BMI}} = e^{0.094} \approx 1.099$$

Each 1-unit increase in BMI is associated with approximately **9.9% higher odds** of diabetes, holding age constant.

**Part (c)** The raw coefficients (0.094 vs. 0.057) are on different scales (BMI units vs. years), so they cannot be compared directly. To compare fairly, use **standardized coefficients** (multiply each $\beta$ by the standard deviation of that predictor) or compare **odds ratios per standard deviation change**.

**Part (d)**

$$OR = e^{0.094 \times 10} = e^{0.94} \approx 2.56$$

A 10-unit increase in BMI is associated with **2.56 times higher odds** of diabetes, holding age constant.
:::

---



13. A clinical trial follows 8 patients after treatment. Their survival times (months) and event status (1 = died, 0 = censored) are:

```{r}
times  <- c(3, 5, 6, 6, 8, 10, 12, 15)
events <- c(1, 1, 1, 0, 1, 0,  1,  0)

library(survival)
fit <- survfit(Surv(times, events) ~ 1)
plot(fit,
     xlab = "Time (months)",
     ylab = "Survival probability",
     main = "Kaplan-Meier Curve",
     lwd  = 2, col = "steelblue",
     conf.int = TRUE)
```



(a) What is the estimated probability of surviving beyond 10 months?

(b) One patient was censored at month 6. What does this mean, and how does the KM estimator handle it?


::: {.callout-tip collapse="true"}
## Solution

**Part (a)**

| Time | $n_{\text{risk}}$ | $d_i$ | $1 - d_i/n_i$ | $\hat{S}(t)$ |
|------|---------|--------|----------------|--------------|
| 3    | 8       | 1      | 7/8            | 0.875        |
| 5    | 7       | 1      | 6/7            | 0.750        |
| 6    | 6       | 1      | 5/6            | 0.625        |
| 8    | 4       | 1      | 3/4            | 0.469        |
| 12   | 2       | 1      | 1/2            | 0.234        |

At $t = 6$, one patient is censored (exits the risk set before the event at that time), so $n_{\text{risk}} = 6$.

**Part (b)**

$\hat{S}(10) = \hat{S}(8) = 0.469$ (no events between $t = 8$ and $t = 10$; the next event is at $t = 12$).

The estimated probability of surviving beyond 10 months is approximately **46.9%**.

**Part (c)** Censoring at month 6 means the patient left follow-up at that point (e.g., withdrew, was lost to follow-up, or the study ended) without experiencing the event. The KM estimator handles censoring by removing the censored patient from the risk set at their censoring time, without treating them as having died. This uses all available information without biasing the estimate.

**Part (d)**

The step-down curve shows survival probability decreasing at each event time. Tick marks on the curve indicate censored observations.
:::


Formulas:
unpooled SE: $$
\mathrm{SE}\!\left(\bar X_1-\bar X_2\right)
= \sqrt{\frac{s_1^{2}}{n_1}+\frac{s_2^{2}}{n_2}}
$$


degrees of freedom: $$ 
\mathrm{df}
=
\frac{\left(\frac{s_T^2}{n_T}+\frac{s_C^2}{n_C}\right)^2}
{\frac{\left(\frac{s_T^2}{n_T}\right)^2}{n_T-1}
+
\frac{\left(\frac{s_C^2}{n_C}\right)^2}{n_C-1}}.
$$

Goodness of fit test

$$
\chi^{2}=\sum_{i}\frac{(\text{Observed}_{i}-\text{Expected}_{i})^2}{\text{Expected}_{i}}
$$
