---
title: "Plotting the data"
format: html
execute:
  echo: true
  warning: false
  message: false
  resources:
  - Lec2_annotated.pdf
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

<!-- Flex container aligns title and button with text margin -->

::: header-row
<h1 class="page-title">



</h1>

<a href="Lec2_annotated.pdf" class="download-btn" download> ⬇ Download Annotated Slides </a>
:::

## 

We will start with a **scatter plot**.

A scatter plot is a tool for examining the relationship between two numerical variables X and Y. It plots each observed (x,y) pair as a dot on the x-y plane.

```{r}
?DNase  #what is the dataset about?
head(DNase) #shows the first rows
names(DNase) # gives you the name of the variables
plot(DNase$conc, DNase$density,
  ylab = attr(DNase, "labels")$y,
  xlab = paste(attr(DNase, "labels")$x, attr(DNase, "units")$x),
  pch = 3)
```

We can visualize *frequency distributions* with **bar plots** (as well as jittered dotplots), **histograms**, and **boxplots**. Bar plots and histograms display the frequency of values (or grouped values). A boxplot displays the minimum, 1st quartile, median, 3rd quartile, and maximum at once. We can also compare multiple distributions at once.

```{r}
hist(DNase$density, breaks=25, main = "",xlab="Density")
boxplot(density ~ Run, data = DNase)
```

The **shape** of a distribution is often informative. It may be, for example: (a) symmetric and bell-shaped, (b) symmetric but not bell-shaped, (c) right-skewed, (d) left-skewed, (e) exponential, or (f) bimodal.

Base R plotting functions are good for quick exploration, but they have aesthetic limitations. We will use a framework called the **grammar of graphics**, implemented in the package **ggplot2**.
