## Plotting the data

We will start with a **scatter plot**.

A scatter plot is a tool for examining the relationship between two numerical variables X and Y. It plots each observed (x,y) pair as a dot on the x-y plane.


```{r}
?DNase  #what is the dataset about?
head(DNase) #shows the first rows
names(DNase) # gives you the name of the variables
plot(DNase$conc, DNase$density,
  ylab = attr(DNase, "labels")$y,
  xlab = paste(attr(DNase, "labels")$x, attr(DNase, "units")$x),
  pch = 3)
```

We can visualize *frequency distributions* with **bar plots** (as well as jittered dotplots), **histograms**, and **boxplots**. Bar plots and histograms display the frequency of values (or grouped values). A boxplot displays the minimum, 1st quartile, median, 3rd quartile, and maximum at once. We can also compare multiple distributions at once.

```{r}
hist(DNase$density, breaks=25, main = "",xlab="Density")
boxplot(density ~ Run, data = DNase)
```

The **shape** of a distribution is often informative. It may be, for example: (a) symmetric and bell-shaped, (b) symmetric but not bell-shaped, (c) right-skewed, (d) left-skewed, (e) exponential, or (f) bimodal.

Base R plotting functions are good for quick exploration, but they have aesthetic limitations. We will use a framework called the **grammar of graphics**, implemented in the package **ggplot2**.


