---
title: "Week 7 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

\begin{center}
{\huge \textbf{Week 7 Section}}
\end{center}
\vspace{-1em}


```{=html}
<div style="
background-color:#eef7ff;
border:1.5px solid #b7d7f0;
border-radius:14px;
padding:20px;
margin-top:1em;
">

<h2 style="margin-top:0;">Chi-square tests concept check</h2>

<table style="width:100%; border-collapse: collapse;">
  <tr>
    <th style="border-bottom: 2px solid #c9dff2; text-align:left; padding: 8px;">Feature</th>
    <th style="border-bottom: 2px solid #c9dff2; text-align:left; padding: 8px;">Goodness-of-fit test</th>
    <th style="border-bottom: 2px solid #c9dff2; text-align:left; padding: 8px;">Test of independence</th>
  </tr>

  <tr>
    <td style="padding: 8px;"><strong>Number of variables</strong></td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">1 categorical variable</span>
      </details>
    </td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">2 categorical variables</span>
      </details>
    </td>
  </tr>

  <tr>
    <td style="padding: 8px;"><strong>Main question</strong></td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">Do category proportions match a hypothesized distribution?</span>
      </details>
    </td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">Are two categorical variables associated?<br>
        Or: Do multiple groups share the same underlying distribution?</span>
      </details>
    </td>
  </tr>

  <tr>
    <td style="padding: 8px;"><strong>Null hypothesis</strong></td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">The observed distribution follows the claimed proportions.</span>
      </details>
    </td>
    <td style="padding: 8px;">
      <details>
        <summary>Answer</summary>
        <span style="color:#1f5aa6;">The two variables are independent.</span>
      </details>
    </td>
  </tr>
</table>

</div>
```

# Example: Campus café visited by student majors
# Part 1: Goodness-of-fit test

Question:

> Do Statistics, Biology, and CS students go to Voyager equally often?

Observed counts:

| Major | Count |
|---|---:|
| Statistics | 39 |
| Biology | 25 |
| CS | 46 |

## Step 1: Write down the hypotheses

Null hypothesis:

**Your answer**

Alternative hypothesis:

**Your answer**


## Step 2: Write down the test statistic

For a goodness-of-fit test,

$$\chi^2=\sum_{i=1}^k\frac{(\text{Observed}_i-\text{Expected}_i)^2}{\text{Expected}_i}$$

with degrees of freedom

$$df = k-1$$

where:

- $\text{Observed}_i$: observed counts
- $\text{Expected}_i$ expected counts under $H_0$


## Step 3: Identify quantities and compute test statistic

If the null hypothesis says all $k$ categories are equally likely, **how should we compute $\text{Expected}_i$**?

```{=html}
<div style="
background-color:#eef7ff;
border:1.5px solid #b7d7f0;
border-radius:14px;
padding:20px;
margin-top:1em;
margin-bottom: 28px;
">

<h3 style="margin-top:0; margin-bottom:22px;">
Observed vs. expected counts
</h3>

<details style="margin-bottom:14px;">
  <summary style="font-weight:600; cursor:pointer;">
    Observed counts
  </summary>

  <ul style="padding-top:14px; margin:0;">
    <li>
      <span style="color:#1f5aa6; font-weight:600;">
      Statistics: 39
      </span>
    </li>

    <li>
      <span style="color:#1f5aa6; font-weight:600;">
      Biology: 25
      </span>
    </li>

    <li>
      <span style="color:#1f5aa6; font-weight:600;">
      CS: 46
      </span>
    </li>
  </ul>
</details>

<details style="margin-bottom:0px;">
  <summary style="font-weight:600; cursor:pointer;">
    Expected counts
  </summary>

  <ul style="padding-top:14px; margin:0 0 0 0;">
    <li>
      <span style="color:#9b3d00; font-weight:600;">
      Statistics: 36.67
      </span>
    </li>

    <li>
      <span style="color:#9b3d00; font-weight:600;">
      Biology: 36.67
      </span>
    </li>

    <li>
      <span style="color:#9b3d00; font-weight:600;">
      CS: 36.67
      </span>
    </li>
  </ul>
</details>

</div>
```

```{r, echo=FALSE}
obs <- c(
  Statistics = 39,
  Biology = 25,
  CS = 46
)
n <- sum(obs)
expected <- rep(n / 3, 3)
names(expected) <- names(obs)

# Compute the chi-square statistic:
chi_square <- sum((obs - expected)^2 / expected)

# Degrees of freedom:
df <- length(obs) - 1
```


- Chi-square statistic: **Your answer**
- Degrees of freedom: **Your answer**

## Step 4: Compute the p-value

- The p-value is the probability, assuming the null hypothesis is true, of observing a test statistic at least as **extreme** as the one observed.

- In a goodness-of-fit test, this means observing counts that **deviate from the claimed distribution** at least as much as the observed data.

**How do we translate the p-value idea into a probability formula?**

```{r, echo = FALSE}
p_value <- pchisq(
  chi_square,
  df = df,
  lower.tail = FALSE
)
```


### Chi-square reference distribution

```{r, echo=FALSE}
library(ggplot2)

x <- seq(0, 20, length.out = 1000)

plot_df <- data.frame(
  x = x,
  density = dchisq(x, df = df)
)

ggplot(plot_df, aes(x = x, y = density)) +

  geom_line(
    linewidth = 1.2,
    color = "steelblue"
  ) +

  geom_area(
    data = subset(plot_df, x >= chi_square),
    aes(x = x, y = density),
    alpha = 0.35,
    fill = "orange"
  ) +

  geom_vline(
    xintercept = chi_square,
    linetype = "dashed",
    linewidth = 1,
    color = "firebrick"
  ) +

  annotate(
    "text",
    x = chi_square,
    y = max(plot_df$density) * 0.85,
    label = "observed",
    color = "firebrick",
    hjust = -0.05
  ) +

  labs(
    title = paste0(
      "Goodness-of-fit test, df = ",
      df
    ),
    x = expression(chi^2),
    y = "Density"
  ) +

  theme_minimal()
```

## Step 5: Make a decision

- If the p-value $<\alpha$, reject $H_0$
- Otherwise, fail to reject $H_0$

Interpretation:

A small p-value suggests that Statistics, Biology, and CS students are not equally represented at Coupa.

# Part 2: Chi-square test for independence

Question:

> Does a student’s major affect which café they go to?

Observed counts:

| Café | Statistics | Biology | CS |
|---|---:|---:|---:|
| Voyager | 39 | 25 | 46 |
| Bytes | 30 | 42 | 28 |
| Coupa | 24 | 29 | 37 |


## Step 1: Write down the hypotheses

Null hypothesis:

**Your answer**

Alternative hypothesis:

**Your answer**

## Step 2: Write down the test statistic

For a contingency table, we use the same goodness-of-fit idea, but apply it across **all cells** of the table. You can think of this as:

- compute the goodness-of-fit contribution within each row
- then sum those contributions across all rows

$$\chi^2=\sum_{i=1}^r\sum_{j=1}^c\frac{(\text{Observed}_{ij}-\text{Expected}_{ij})^2}{\text{Expected}_{ij}}$$

with degrees of freedom: $$df = (r-1)(c-1)$$

If the row distributions are very different from the shared distribution expected under independence, the chi-square statistic becomes large.

## Step 3: Identify quantities and compute

Under the null hypothesis of independence:

- the row category should not affect the column category
- every row should follow the same overall column distribution: $\frac{\text{column total}}{\text{grand total}}$

Therefore expected counts: $$\text{Expected}_{ij}=\frac{(\text{row total})\times(\text{column total})}{\text{grand total}}$$


```{=html}
<div style="
background-color:#eef7ff;
border:1.5px solid #b7d7f0;
border-radius:14px;
padding:20px;
margin-top:1em;
margin-bottom:28px;
">

<h3 style="
margin-top:0;
margin-bottom:26px;
">
Observed vs. expected contingency tables
</h3>

<details style="margin-bottom:18px;">
  <summary style="
  font-weight:600;
  cursor:pointer;
  ">
    Observed counts
  </summary>

  <table style="
  width:100%;
  border-collapse:collapse;
  margin-top:18px;
  ">

    <tr>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;"></th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        Statistics
      </th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        Biology
      </th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        CS
      </th>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Voyager</strong></td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">39</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">25</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">46</td>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Bytes</strong></td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">30</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">42</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">28</td>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Coupa</strong></td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">24</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">29</td>
      <td style="padding:8px; color:#1f5aa6; font-weight:600;">37</td>
    </tr>

  </table>
</details>

<details style="margin-bottom:0;">
  <summary style="
  font-weight:600;
  cursor:pointer;
  ">
    Expected counts
  </summary>

  <table style="
  width:100%;
  border-collapse:collapse;
  margin-top:14px;
  margin-bottom:0;
  ">

    <tr>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;"></th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        Statistics
      </th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        Biology
      </th>
      <th style="padding:8px; border-bottom:2px solid #c9dff2;">
        CS
      </th>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Voyager</strong></td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">34.1</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">35.2</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">40.7</td>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Bytes</strong></td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">31.0</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">32.0</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">37.0</td>
    </tr>

    <tr>
      <td style="padding:8px;"><strong>Coupa</strong></td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">27.9</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">28.8</td>
      <td style="padding:8px; color:#9b3d00; font-weight:600;">33.3</td>
    </tr>

  </table>
</details>

</div>
```

```{r,echo=FALSE}
obs_table <- matrix(
  c(
    39, 25, 46,
    30, 42, 28,
    24, 29, 37
  ),
  nrow = 3,
  byrow = TRUE
)

rownames(obs_table) <- c(
  "Voyager",
  "Bytes",
  "Coupa"
)

colnames(obs_table) <- c(
  "Statistics",
  "Biology",
  "CS"
)

row_totals <- rowSums(obs_table)
col_totals <- colSums(obs_table)
grand_total <- sum(obs_table)

# Expected counts:
expected <- outer(
  row_totals,
  col_totals
) / grand_total
```



```{r, echo = FALSE}
# Compute the chi-square statistic:
chi_square <- sum(
  (obs_table - expected)^2 / expected
)

#Degrees of freedom:
r <- nrow(obs_table)
c <- ncol(obs_table)
df <- (r - 1) * (c - 1)
```

- Chi-square statistic: **Your answer**
- Degrees of freedom: **Your answer**


## Step 4: Compute the p-value

- The p-value is the probability, assuming the null hypothesis is true, of observing a test statistic at least as **extreme** as the one observed.

- In a test of independence, this means observing a contingency table that **deviates from the independence pattern** at least as much as the observed table:

$$p\text{-value}=\mathbb{P}\!\left[\chi^2_{df}\ge\chi^2_{\mathrm{obs}}\right]$$

```{r, echo = FALSE}
p_value <- pchisq(
  chi_square,
  df = df,
  lower.tail = FALSE
)
```


# Chi-square reference distribution
```{r, echo=FALSE}
x <- seq(0, 30, length.out = 1000)

plot_df <- data.frame(
  x = x,
  density = dchisq(x, df = df)
)

ggplot(plot_df, aes(x = x, y = density)) +

  geom_line(
    linewidth = 1.2,
    color = "steelblue"
  ) +

  geom_area(
    data = subset(plot_df, x >= chi_square),
    aes(x = x, y = density),
    alpha = 0.35,
    fill = "orange"
  ) +

  geom_vline(
    xintercept = chi_square,
    linetype = "dashed",
    linewidth = 1,
    color = "firebrick"
  ) +

  annotate(
    "text",
    x = chi_square,
    y = max(plot_df$density) * 0.85,
    label = paste0(
      "observed = ",
      round(chi_square, 2)
    ),
    color = "firebrick",
    hjust = -0.05
  )  +

  labs(
    subtitle = paste0(
      "Test of independence, df = ",
      df
    ),
    x = expression(chi^2),
    y = "Density"
  ) +

  theme_minimal()
```

## Step 5: Make a decision

- If the p-value $<\alpha$, reject $H_0$
- Otherwise, fail to reject $H_0$

Interpretation:

A small p-value suggests that café choice differs across Statistics, Biology, and CS students.