---
title: "Week 8 Section"
author: "Stats 141, Spring 2026"
format: pdf
execute:
  echo: true
  warning: false
  message: false
---

Acknowledgement: Much of this material is borrowed from chapter 3 of the book Introduction to Statistical Learning with R (ISLR).

We will work with the `Advertising` dataset from ISLR. It records `sales` (in thousands of units) alongside advertising budgets (in thousands of dollars) for `TV`, `radio`, and `newspaper` across 200 markets.

```{r}
Advertising <- read.csv("https://www.statlearning.com/s/Advertising.csv")[,-1]
head(Advertising)
```

The dataset raises several natural questions:

- Is there a relationship between advertising budget and sales?
If the evidence is weak, then one might argue that no money should be
spent on advertising!

- Are all three media (TV, radio, and newspaper) associated with
sales, or are just one or two of them? To answer this
question, we must find a way to separate out the individual contribution of each medium to sales when we have spent money on all three
media.

- For every dollar spent on advertising in a particular medium, by
what amount will sales increase? How accurately can we predict this?

- How adequate is the linear model for the relationship between sales and different media?

---

# Problem 1: Exploring correlations

Before fitting any model, we should always look at the data. 

```{r fig.height = 4.5, fig.width = 5.5, fig.align='center'}
pairs(Advertising,
      col = "steelblue", pch = 19, cex = 0.4)
```

**(a)** Compute the correlation between each advertising channel and `sales`. Which of these is most strongly linearly associated with sales?

```{r}
# use cor()
```

`TV` has the strongest linear association with sales ($r \approx 0.78$), followed by `radio` ($r \approx 0.58$). `newspaper` shows a weaker correlation ($r \approx 0.23$).


**(b)** The correlation coefficient is always between $-1$ and $1$ and is defined as
$$r = \frac{1}{n-1} \sum_{i=1}^n \left(\frac{x_i - \bar{x}}{s_x}\right)\left(\frac{y_i - \bar{y}}{s_y}\right).$$
What does the sign tell you? What does the magnitude tell you?

The sign gives the direction of the association, it answers whether an increase in $x$ is associated with an increase or decrease in $y$. The magnitude reflects how tightly the points cluster around a line: $|r| = 1$ means perfect linear association, $r = 0$ means no linear association.

---

# Problem 2: Simple linear regression

## Fitting the model

We regress `sales` on `TV` advertising budget alone.

```{r}
# Use lm and summary
# model_tv <- 
```

**(a)** Write out the fitted model. Interpret the slope.

$$\widehat{\text{sales}} = 7.033 + 0.0475 \times \text{TV}$$

Recall that `sales` is in thousands of units, and `TV` budget is in \$1,000. Thus, the model fitted above tells us that: For every additional \$1,000 spent on TV advertising, sales increase by about 47.5 units. The intercept (7,033 units) is the predicted sales when TV budget is zero.

```{r fig.height = 4, fig.width = 6, fig.align='center'}
plot(Advertising$TV, Advertising$sales,
     col = "steelblue", pch = 19, cex = 0.6,
     xlab = "TV budget ($000s)", ylab = "Sales (000s)",
     main = "Sales vs. TV advertising")
# abline(model_tv, col = "tomato", lwd = 2)
```

## Deriving the OLS estimates

The least squares estimates minimize the residual sum of squares $\text{RSS} = \sum_{i=1}^n (y_i - \hat{y}_i)^2$. Using calculus, this yields the following closed forms:
$$\hat{\beta}_1 = \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n (x_i - \bar{x})^2} = \text{cor}(x,y)\times \frac{\text{sd(y)}}{sd(x)}, \qquad \hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}.$$

**(b)** Verify these manually.

```{r}
# x <- 
# y <- 
# 
# beta1_hat <- 
# beta0_hat <- 
# 
# c(intercept = beta0_hat, slope = beta1_hat)
```

These should match the `lm` output exactly.

## Inference on the slope

Standard errors quantify how much $\hat\beta_1$ would vary across repeated samples. A 95% confidence interval for the true slope $\beta_1$ is

$$\hat\beta_1 \pm t_{0.975,\, n-2} \cdot \text{SE}(\hat\beta_1).$$

```{r}
# confint(model_tv)
```

We are 95% confident that each additional \$1,000 in TV budget raises sales by between 42.5 and 52.4 units.

To test $H_0: \beta_1 = 0$ (no relationship), we use the $t$-statistic

$$t = \frac{\hat\beta_1}{\text{SE}(\hat\beta_1)},$$

which follows a $t_{n-2}$ distribution under $H_0$. The p-value from `summary(model_tv)` is essentially zero. This suggests that *there is strong evidence of a relationship between TV spending and sales*.

## Model fit / goodness-of-fit

Two standard measures:

- **RSE** (residual std err): prediction error in the units of the outcome, $\text{RSE} = \sqrt{\text{RSS}/(n-2)}$.
- **$R^2$**: proportion of variance in the outcome (`sales`) explained by the predictor (`TV`), $R^2 = 1 - \text{RSS}/\text{TSS}$.

```{r}
# n   <- nrow(Advertising)
# rss <- sum(residuals(model_tv)^2)
# tss <- sum((y - mean(y))^2)
# 
# c(RSE = sqrt(rss / (n - 2)), R2 = 1 - rss / tss)
```

We could also read off these numbers from the summary:

```{r}
# summary(model_tv)
```

TV budget explains about 61% of the variation in sales. Predictions are off by roughly 3,260 units on average. There is still substantial unexplained variation---maybe including other predictors can help?

---

# Problem 3: Multiple linear regression

## Why not just fit three separate models?

We could regress `sales` on each predictor in turn. The problem is that predictors are correlated, so each simple regression confounds the effect of one predictor with the others. Multiple regression estimates each coefficient **holding the others fixed**, separating out the individual contributions.

## Fitting the model

```{r}
# model_all <-
# summary(model_all)
```

## Which media matter most?

Looking at individual p-values in the multiple regression output: `TV` and `radio` are strongly significant, while `newspaper` has p-value $\approx 0.86$. This suggests that **TV and radio are associated with sales and newspaper is not**---*once we control for the other two*.


**(a)** Compute confidence intervals for the effects of each medium (i.e., confidence intervals for all the coefficients in the multiple regression).

```{r}
# use confint
```

Holding the other budgets fixed:

- Each additional \$1,000 in **TV** raises sales by roughly 46–49 units.
- Each additional \$1,000 in **radio** raises sales by roughly 172–207 units.
- **Newspaper** spending has no detectable effect (CI includes zero).

**(b)** Compare the `newspaper` coefficient in simple versus multiple regression.

```{r}
# model_news <- 
# c(simple = coef(model_news)["newspaper"],
#   multiple = coef(model_all)["newspaper"])
```

In simple regression, `newspaper` looks like it has a positive effect. In multiple regression it vanishes. Why? Because `newspaper` and `radio` spending are correlated across markets. The simple regression picks up the `radio` effect and mistakenly attributes it to `newspaper`. This is an example of *confounding*.

## Model fit / goodness-of-fit

The process will similar to earlier, now for the multiple regression model.

```{r}
# rss <- sum(residuals(model_all)^2)
# m <- 3 # number of predictors
# 
# c(RSE = sqrt(rss / (n - m - 1)), # note, when m = 1, this is n-2
#   R2  = 1 - rss / tss)
```

We could also do it directly from the `lm` summary:

```{r}
# use lm and summary
```

Including all three predictors, $R^2$ jumps from 0.61 (TV alone) to 0.90, and RSE drops from 3.26 to 1.69. Almost all of this gain comes from adding `radio`; `newspaper` contributes essentially nothing, as we can check:

```{r}
# use lm and summary
```

Side note (probably out of scope): $R^2$ always increases when you add predictors, even irrelevant ones (it can only decrease RSS on the *training data*). When you have many predictors, this can lead to 'overfitting'. This is why we look at measures like adjusted $R^2$, AIC, or BIC when comparing models of different sizes.


## Testing all predictors simultaneously: the F-test

In multiple regression with $m$ predictors, the global null hypothesis is $H_0: \beta_1 = \cdots = \beta_m = 0$. We test it with the F-statistic:

$$F = \frac{(\text{TSS} - \text{RSS})/m}{\text{RSS}/(n - m - 1)}.$$

Under $H_0$, $F \approx 1$. So, a large $F$ is evidence against $H_0$. In our `lm` output, $F = 570.3$ with a p-value near zero. This suggest that *at least one predictor is associated with sales*.

---
