# Activity 18 

Name:



1. For each example, state whether or not the censoring mechanism is independent. Justify your answer.

(a) In a study of disease relapse, due to a careless research scientist, all patients whose phone numbers begin with the number "2" are lost to follow up.

(b) In a study of longevity, a formatting error causes all patient ages that exceed 99 years to be lost (we know that those patients are more than 99 but we do not know their exact ages).

(c) Hospital A conducts a study of longevity. However, very sick patients tend to be transferred to Hospital B, and are lost to follow up.


2. The R package survival includes a dataset lung that contains survival times for patients with advanced lung cancer. Key variables:

time: follow-up time (days)
status: 1=censored, 2=dead (event) 
Common covariates:
age (years), sex (1=male, 2=female), ph.ecog (ECOG performance status), plus others.

```{r}
#| warning: false
library(survival)
data(lung)
```
(a) What is the Kaplan–Meier estimate for the whole sample?


```{r}
km_all <- survfit(Surv(time, status) ~ 1, data = lung)
plot(km_all, xlab="Days", ylab="Survival", main="KM: All patients")
abline(h=0.5)
abline(v=310)

```
(b) What is the median survival time (overall), that is the time when $\hat S(t)=0.5$?

```{r}
summary(km_all)$table["median"]
```

(c) Do survival curves differ by sex (unadjusted)?

```{r}
km_sex <- survfit(Surv(time, status) ~ sex, data = lung)
plot(km_sex, col=c("red","blue"), lty=1:2, xlab="Days", ylab="Survival")
legend("topright", legend=c("Male (1)","Female (2)"),
       col=c("red","blue"), lty=1:2, bty="n")
```

(d) Is the difference by sex statistically significant (logrank test)?

```{r}
survdiff(Surv(time, status) ~ sex, data = lung)
```
a small p-value suggests different survival functions between sexes.

e) How do we fit a Cox proportional hazards model with key predictors?

```{r}
cox1 <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data = lung)
summary(cox1)
```
f) How do we report hazard ratios with 95% CIs?

```{r}
#Exponentiate coefficients and confidence intervals.
hr  <- exp(coef(cox1))
ci  <- exp(confint(cox1))
out <- cbind(HR = hr, `2.5%` = ci[,1], `97.5%` = ci[,2])
round(out, 3)
```
g) How do we interpret coefficients?
Answer:

If HR(age)=1.02, each additional year of age multiplies hazard by 1.02 (~2% higher hazard), holding other variables fixed.
If HR(sex)=0.70 (female vs male, depending on coding), females have 30% lower hazard, adjusted.
