---
title: "Week 5 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
resources:
  - Week5_section_incomplete.pdf
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

::: header-row
<h1 class="page-title">


</h1>

<a href="Week5_section_incomplete.pdf" class="download-btn" download> ⬇
Download Worksheet </a>
:::

# Learning goals

By the end of this section, you should be able to:

1.  Construct one-sample confidence intervals using both the $Z$ and $t$
    methods.
2.  Explain why the $Z$ and $t$ intervals are slightly different.
3.  Conduct a two-sample mean hypothesis test in two equivalent ways:
    -   by constructing a confidence interval for the mean difference;
    -   by computing a test statistic and p-value.

We encourage you to work out the following problems with pen and paper,
and treat R as a calculator/look-up table. Hence, most commands are
provided. :)

# Problem 1: Confidence intervals from a butterfly field study

A biology team is studying a certain butterfly species. They collected a
random sample of butterflies. Each butterfly is measured from wingtip to
wingtip, then released.

The wingspans, measured in millimeters, are:

```{r}
wingspan <- c(48.6, 52.1, 50.4, 49.7, 53.0, 51.2, 54.1, 50.8,
              52.7, 49.9, 51.6, 53.4, 50.2, 48.9, 52.3, 51.0)

data.frame(butterfly = 1:length(wingspan), wingspan_mm = wingspan)
```

The researchers want a 95% confidence interval for the true mean
wingspan $\mu$ of butterflies in this area.

## Step 1.a: Z confidence interval

The $Z$ interval is appropriate when the population standard deviation
$\sigma$ is known.

Write down a 95% confidence interval using the $Z$ method, pretending
the population standard deviation is known and equal to $\sigma = 1.6$
mm. Identify which summary statistics to use.

**Your answer:**

## Step 1.b: t confidence interval

In most real field studies, $\sigma$ is unknown, so we use the sample
standard deviation $s$ and a $t$ critical value.

Write down a 95% confidence interval using the $t$ method, treating
$\sigma$ as unknown. Identify which summary statistics to use.

**Your answer:**

## Step 2: Summary statistics

Compute the sample mean and sample standard deviation.

```{r}
n <- length(wingspan)
xbar <- mean(wingspan)
s <- sd(wingspan)
```

For a Z-confidence interval, we need

```{r}
sigma <- 1.6
se_z <- sigma / sqrt(n) # SE is SD/sqrt(n)
```

For a t-confidence interval, we need

```{r}
df <- n - 1
se_t <- s / sqrt(n)
```

## Step 3: Identify confidence level and look up critical-value

For a two-sided 95% interval, $\alpha = 0.05$:

```{r}
lookup_table <- data.frame(
  method = c("Z", "t"),
  probability_to_left = c(0.975, 0.975),
  degrees_of_freedom = c(NA, 15),
  critical_value = c(qnorm(0.975), qt(0.975, 15))
)

lookup_table
```

A traditional table lookup gives approximately:

-   $z_{0.975} = 1.96$;
-   $t_{0.975,15} = 2.131$.

## Step 4: Plug in numbers

For a 95% Z-confidence interval, we use $z_{0.975}$.

```{r}
ci_z <- c(xbar - 1.96 * se_z,
          xbar + 1.96 * se_z)

ci_z
```

For a 95% t-confidence interval, we use $t_{0.975, 15}$.

```{r}
ci_t <- c(xbar - 2.131 * se_t,
          xbar + 2.131 * se_t)

ci_t
```

## Why are the intervals different?

### Plot the Z and t distributions

```{r}
curve(dnorm(x), from = -4, to = 4,
      col = "steelblue", lwd = 2,
      main = "Standard Normal vs t Distribution",
      xlab = "x", ylab = "Density")

curve(dt(x, df = 15), from = -4, to = 4,
      col = "tomato", lwd = 2, add = TRUE)

abline(v = c(-1.96, 1.96), col = "steelblue", lty = 2)
abline(v = c(-2.131, 2.131), col = "tomato", lty = 2)

legend("topright",
       legend = c("Z: standard normal", paste0("t: df = ", 15)),
       col = c("steelblue", "tomato"),
       lwd = 2,
       bty = "n")
```

**Your answer:**

Hint: Which confidence interval is wider? Can you explain why?

# Problem 2: Two-sample test from two butterfly habitats

The same research team now compares butterflies from two habitats A and
B. The team wonders whether butterflies from the two habitats have
different mean wingspans. Again, they want to use significance level
$\alpha = 0.05$.

The measured wingspans are:

```{r}
habitat_A <- c(48.6, 52.1, 50.4, 49.7, 53.0, 51.2, 54.1, 50.8,
            52.7, 49.9, 51.6, 53.4, 50.2, 48.9, 52.3, 51.0)

habitat_B <- c(46.8, 49.1, 47.5, 48.3, 50.0, 46.9, 49.7, 47.8,
            48.6, 45.9, 50.4, 47.2)

butterflies <- data.frame(
  habitat = c(rep("A", length(habitat_A)), rep("B", length(habitat_B))),
  wingspan_mm = c(habitat_A, habitat_B)
)

butterflies
```

The purpose of a hypothesis test is to test a claim and determine if
there is sufficient evidence in favor of that claim. Hypothesis testing
always follows a standard procedure, we can break it down to the
following **5 steps**.

## Step 1: Write down the hypothesis

What is the null and alternative hypothesis?

**Your answer:**

## Step 2.a: Construct confidence intervals

One way is to use a confidence interval for $\mu_{A} - \mu_{B}$ to test
the null hypothesis. Write down a 95% confidence interval using the $t$
method, treating standard deviation as unknown. Identify which summary
statistics to use.

**Your answer:**

## Step 2.b: Compute test statistics

Alternatively, we can use a p-value to test the null hypothesis. Write
down the test statistics under the null hypothesis. Identify which
summary statistics to use.

**Your answer:**

## Step 3: Summary statistics

First compute the sample means for the two habitats.

```{r}
xbar1 <- mean(habitat_A)
xbar2 <- mean(habitat_B)
```

The observed mean difference is

```{r}
diff_means <- xbar1 - xbar2
diff_means
```

Note that degree of freedom and standard error are more complicated, we
have introduced two methods to estimate them in class -- Welch's
unpooled standard error, and pooled standard error. Welch's method is
usually the safer default because it does not require equal population
variances. The pooled test can be reasonable when the two groups are
believed to have similar variances, either from scientific knowledge or
from the data.

In this exercise, we focus on using the Welch's unpooled standard error,

$$SE_{\text{Welch}} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}.$$

The Welch degrees of freedom are approximated by

$$df = \frac{\left(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}\right)^2}
{\frac{(s_1^2/n_1)^2}{n_1-1} + \frac{(s_2^2/n_2)^2}{n_2-1}}.$$ To
compute the standard error, we also need sample sizes and standard
deviations from the data:

```{r}
n1 <- length(habitat_A)
n2 <- length(habitat_B)

s1 <- sd(habitat_A)
s2 <- sd(habitat_B)

data.frame(
  habitat = c("habitat_A", "habitat_B"),
  n = c(n1, n2),
  mean = c(xbar1, xbar2),
  sd = c(s1, s2)
)
```

Then we can put everything together and compute the standard error:

```{r}
se_welch <- sqrt(s1^2 / n1 + s2^2 / n2)

df_welch <- (s1^2 / n1 + s2^2 / n2)^2 /
  ((s1^2 / n1)^2 / (n1 - 1) + (s2^2 / n2)^2 / (n2 - 1))

tcrit_welch <- qt(0.975, df = df_welch)
```

## Step 4: Plug in numbers

For our 95% confidence interval, we have

```{r}
ci_welch <- c(diff_means - tcrit_welch * se_welch,
              diff_means + tcrit_welch * se_welch)
ci_welch # confidence interval
```

For our t-test statistics, we have

```{r}
t_welch <- diff_means / se_welch
p_welch <- 2 * (1 - pt(abs(t_welch), df = df_welch))

t_welch # test statistics
p_welch # p-val
```

(Alternative) Using R's built-in function:

```{r}
t.test(habitat_A, habitat_B, alternative = "two.sided", var.equal = FALSE)
```

## Step 5: Make a decision

Confidence interval interpretation:

**Your answer:**

p-value interpretation:

**Your answer:**

## Conclusion

Both approaches answer the same question: is the mean wingspan different
between the two habitats?

The confidence-interval approach and the p-value approach agree:

-   If the 95% confidence interval excludes 0, then the two-sided
    p-value is below 0.05.
-   If the 95% confidence interval includes 0, then the two-sided
    p-value is above 0.05.
