# Activity 13 {#sec-week2}

Name:

1.Substances to be tested for cancer-causing potential are often painted on the skin of mice. The question arose whether mice might get an additional dose of the substance by licking or biting their cagemates. To answer this question, the compound benzo(a)pyrene was applied to the backs of 10 mice: Five were individually housed
and 5 were group-housed in a single cage. After 48 hours,
the concentration of the compound in the stomach tissue
of each mouse was determined. The results (nmol/gm)
were as follows:

```{r}
single<-c(3.3,2.4,2.5,3.3,2.4)
group<-c(3.9,4.1,4.8,3.9,3.4)

```

Use a Wilcoxon-Mann-Whitney test to compare the
two distributions at $\alpha = 0.01$. Let the alternative 
hypothesis be that benzo(a)pyrene concentrations
tend to be higher in group-housed mice than in
singly housed mice.

2. In an experiment to compare two diets for fattening beef steers, nine pairs of animals were chosen from the herd; members of each pair were matched as closely as possible with respect to hereditary factors. The members of each pair were randomly allocated, one to each diet. The following data shows the weight gains (lb) of the animals over a 140-day test period on diet 1 (Y1) and
on diet 2 (Y2).

```{r}
diet1<-c(596,422,524,454,538,552,478,564,556)
diet2<-c(498,460,468,458,530,482,528,598,456)
```

(a) Calculate the standard error of the mean difference.

(b) Test for a difference between the diets using a paired t-test at
$\alpha = 0.10$. Use a nondirectional alternative.

(c) Construct a 90% confidence interval for $\mu_D$.

(d) Interpret the confidence interval from part (c) in the
context of this setting.

