---
#title: "Week 2 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
resources:
  - Week2_section_download.qmd
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

<!-- Flex container aligns title and button with text margin -->

::: header-row
<h1 class="page-title">

Week 2 Section

</h1>

<a href="Week2_section_download.qmd" class="download-btn" download> ⬇ Download Starter File </a>
:::


# Part 1: Correlation vs Causation

## Example 1: Are students more likely to default on a credit card?

### Background (`ISLR::Default`)

The `Default` dataset (from the `ISLR` package) records 10,000 credit card customers. We want to understand what predicts default.

**Exposure**: `student` -- is the customer a student?\
**Outcome**: `default` -- did the customer default?\
**Possible confounders**: `balance` -- the customer's average credit card balance

```{r}
# Load the package ISLR
# Load the data Default from ISLR package 
# Check head of the data and structure
```

### Are students more likely to default than non-students?

```{r}
# Compute mean default rates for students vs non-students
```

Should get: Students have a 4.3% chance of defaulting, whereas for non-students the same chance is 2.9%.

Next, we dive deeper, trying to understand whether balance is a confounder here.

### Check 1: Are students carrying higher balances?

```{r}
# Compare mean balances for students and non-students
```

### Check 2: Is balance associated with default?

```{r}
# Compute correlation between balance and default
```

### Check 3: Default rates within balance groups

```{r}
# First, compute quantiles of the variable balance

# Next, 
# - figure out which balance group each person belongs to, 
# - add it the original data as an additional feature/column

# Hint:
# Default$balance_group <- cut(Default$balance, breaks = quantiles from above)
```

```{r}
# Check that default rate rises steeply with balance
```

If balance is really the confounder, then at the *same* balance level, students should no longer look riskier.

```{r}
# tapply(Default$default_num,
#        list(balance = Default$balance_group, student = Default$student),
#        mean)
```

Within every single balance quartile, students default at a **lower** rate than non-students -- the opposite of the marginal comparison! This **reversal** is an example of **Simpson's paradox** (Google it to know more).

## Example 2: Do Manual Transmissions Get Better Mileage? (For illustration, will skip in section)

### Background (`mtcars`)

The `mtcars` dataset (built into R) records fuel economy and physical characteristics for 32 cars from a 1974 Motor Trend road test.

**Exposure**: `am` -- transmission type (0 = automatic, 1 = manual)\
**Outcome**: `mpg` -- fuel efficiency (miles per gallon)\
**Possible confounder**: `wt` -- car weight (in thousands of pounds)

```{r}
# load the data mtcars
```

### Do manual cars give more mpg?

```{r}
# Compute the mean mpg for each group (auto/manual) and compute the difference.
```

Manual transmission cars average **24.4 mpg** vs **17.1 mpg** for automatics---a gap of **7.3 mpg**, which is enormous (about 43% better). If you only saw this number, you might conclude that choosing a manual gearbox is the key to fuel savings.

### Are manual and automatic cars different in weight?

```{r}
# Compute the mean weight for each group (auto/manual).
```

Manual cars average **2.41 thousand lbs** vs **3.77 thousand lbs** for automatics -- a difference of **1.36 thousand lbs**, or over half a ton. In earlier days, manual transmissions were common in small, sporty cars while automatics dominated larger, heavier vehicles.

### Is weight associated with mpg?

```{r}
# Compute correlation
```

Heavier cars get much worse mileage. This is intuitive: more mass requires more energy.

### mpg of automatic vs manual cars within weight groups

```{r}
# First, compute quantiles of the weights
qs <- quantile(mtcars$wt, probs = c(0, 1/3, 2/3, 1))

# Then, figure out which group each car belongs to, add it the data as a column
mtcars$wt_group <- cut(mtcars$wt, 
                       breaks = qs,
                       labels = c("Light", "Medium", "Heavy")
                       )
```

```{r}
# Compare the mpg of automatic versus manual in each weight group
# tapply(mtcars$mpg, list(wt_group = mtcars$wt_group, am = mtcars$am), mean)
```

### Summary

In this dataset,

-   Manual cars are much lighter.
-   Lighter cars get much better mileage.
-   Within similar weight classes, the manual–automatic mpg gap is smaller.

# Part 2: Probability

The **probability** of an event is the proportion of times it occurs in a very large number of identical, independent repetitions of the same random experiment. We can use random number generators in R to *simulate* these repetitions and see this in action.

## Simulating a Coin Flip

Suppose we flip a fair coin. The probability of getting Heads should be 0.5. Let's verify this by simulation.

```{r}
# Simulate 10 coin flips: 1 = Heads, 0 = Tails

# Print the proportion of Heads in those 10 flips
```

```{r}
# Repeat the above for 1,000 flips
```

The frequency gets closer and closer to the probability (0.5) as we increase the number of flips.

## Seeds for reproducibility

```{r}
# Scenario 1:
sample(c(0, 1), size = 10, replace = TRUE)
sample(c(0, 1), size = 10, replace = TRUE)

# Scenario 2:
set.seed(123)
sample(c(0, 1), size = 10, replace = TRUE)

set.seed(123)
sample(c(0, 1), size = 10, replace = TRUE)
```

## Simulating a biased coin

Now suppose the coin is biased, with ( P(H) = 0.7, P(T) = 0.3. )

```{r}
biased_coin <- sample(c("H", "T"),
                      size = 1000,
                      replace = TRUE,
                      prob = c(0.7, 0.3) # this is the part that's new
                      )
# Compute proportion of Heads and Tails
```

The simulated proportions should be close to 0.7 and 0.3.

## Calculating Probability of an Event: Rolling Two Dice

Now let's simulate rolling two fair six-sided dice and ask: **What is the probability that the sum equals 7? What about 8?**

```{r}
n_rolls <- 10000   # number of times we roll the pair of dice

# Simulate the experiment

# sums <- die1 + die2

# Estimate probability that sum = 7

# Estimate probability that sum = 8
```

We can compare to the **exact** probabilities. There are $6 \times 6 = 36$ equally likely outcomes. The ways to get a sum of 7 are: (1,6), (2,5), (3,4), (4,3), (5,2), (6,1) — that is 6 outcomes. For sum = 8: (2,6), (3,5), (4,4), (5,3), (6,2) — that is 5 outcomes.

```{r}
# Exact probabilities
6/36   # P(sum = 7)
5/36   # P(sum = 8)
```

Our simulated values should be very close! (Because we simulated many identical, independent repetitions of this experiment.)

We can also look at the full distribution of the sum of two dice:

```{r}
# Frequency table of all possible sums
```

```{r}
# Visualize
# barplot(table(sums) / n_rolls,
#        main = "Simulated Probability Distribution of Sum of Two Dice",
#        xlab = "Sum of Two Dice",
#        ylab = "Estimated Probability",
#        col = "steelblue")

# Add a horizontal line at the exact P(sum=7) = 6/36 for reference
```

Notice the distribution is **symmetric** around 7 and **triangular** in shape — sums close to 7 are most likely.

## Biased die

Now let the die be biased. Suppose the probabilities are:

-   1 with probability 0.05
-   2 with probability 0.10
-   3 with probability 0.15
-   4 with probability 0.20
-   5 with probability 0.20
-   6 with probability 0.30

```{r}
prob_die <- c(0.05, 0.10, 0.15, 0.20, 0.20, 0.30)
sum(prob_die)
```

```{r}
# Simulate the biased roll as we did above for the biased coin

# biased_rolls <- sample(...)
```

### What is the probability of rolling a 5 or 6?

```{r}
# Compute this using the simulated experiment
```

Exact answer from the probability distribution:

```{r}
# Compute using prob_die
```

### Display the distribution of the biased die

Because the outcomes are discrete, a barplot is often the first choice.

```{r}
# barplot(table(biased_rolls),
#         main = "Biased die outcomes",
#         xlab = "Outcome",
#         ylab = "Count",
#         col = "tomato")
```

We can also draw a histogram. `hist(biased_rolls, ...)`

### Sum of two rolls of the biased die

We can also look at the distribution of the sum of two biased dice. Following code is essentially same as before, this time with `prob = prob_die` in `sample()`.

```{r}
# Simulate this (recycle code from earlier)
```

### Compare it with the same for the fair dice

Reproducing the plot from earlier (no need to run the experiment again!):

```{r}
# barplot(table(sums) / n_rolls,
#         main = "Simulated Probability Distribution of Sum of Two Dice",
#         xlab = "Sum of Two Dice",
#         ylab = "Estimated Probability",
#         col = "steelblue")
```
