---
#title: "Week 4 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
resources:
  - Week4_section_download.qmd
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

<!-- Flex container aligns title and button with text margin -->

::: header-row
<h1 class="page-title">

Week 4 Section

</h1>

<a href="Week4_section_download.qmd" class="download-btn" download> ⬇
Download Starter File </a>
:::

# Week 4

# Part 1: Expectation and Variance

**Problem.** Alice and Bob have a machine that generates random
observations from an unknown distribution. Alice generates one
observation $X$ and draws a square of length $X$. Bob generates two
independent observations $X_1$ and $X_2$ and draws a rectangle of
dimensions $X_1$ by $X_2$. Which of these rectangles (a square is also a
rectangle!) is expected to have the larger area?

**Hint:** $X$, $X_1$ and $X_2$ are independent observations from the
same distribution.

# Part 2: Normal distribution

The normal distribution $X \sim N(\mu, \sigma^2)$ is characterized by
two parameters: the mean $\mu$ (location) and the variance $\sigma^2$
(spread). R provides four key functions for working with it:

| Function | What it does | Analog for Binomial |
|------------------------|------------------------|------------------------|
| `dnorm(x, mean, sd)` | Height of density curve at $x$ | `dbinom` (PMF) |
| `pnorm(q, mean, sd)` | $P(X \leq q)$ — area to the left | `pbinom` |
| `qnorm(p, mean, sd)` | Quantile: value with area $p$ to its left | `qbinom` |
| `rnorm(n, mean, sd)` | Draw $n$ random observations | `rbinom` |

## Visualizing normal curves

The parameter $\mu$ shifts the curve left or right. The parameter
$\sigma$ changes how wide or narrow the curve is.

```{r}
x <- seq(80, 360, length.out = 1000)

plot(x, dnorm(x, mean = 220, sd = 20),
     type = "l", lwd = 2,
     ylim = c(0, 0.022),
     xlab = "x (Cholesterol)", ylab = "Density",
     main = "Normal curves with different means / SDs")
lines(x, dnorm(x, mean = 220, sd = 35), lwd = 2, col = "steelblue")
lines(x, dnorm(x, mean = 250, sd = 35), lwd = 2, col = "tomato")
legend("topright",
       legend = c("N(220, 20^2)", "N(220, 35^2)", "N(250, 35^2)"),
       col = c("black", "steelblue", "tomato"),
       lwd = 2,
       bty = "n")
```

-   The two curves with mean 220 are centered at the same place.
-   The curve with SD 20 is narrower and taller.
-   Changing the mean moves the whole curve.

Important point: for a continuous random variable, **probability is area
under the curve**, not the height of the curve at one point.

## Standardization: going from X to Z

Suppose cholesterol in some population is approximately normal with mean
220 mg/dL and s.d. 20 mg/dL. Write this as $X\sim N(220, 20^2)$.

To standardize, we compute $$
Z = \frac{X - \mu}{\sigma}.
$$ Then $Z \sim N(0,1)$.

This lets us translate values on the original scale to the standard
normal scale.

```{r}
mu <- 220
sigma <- 20

# Convert 260 mg/dL to a z-score
```

So 260 mg/dL is \_\_ standard deviations above the mean (assuming that
the mean is $220$ and the s.d. is $20$).

## Calculating tail probabilities

For a normal distribution, approximately:

-   68% of values lie within 1-SD away from the mean
-   95% of values lie within 2-SD away from the mean
-   99% of values lie within 3-SD away from the mean

### Example 1: Probability in an interval

Compute $P(180 \leq X \leq 260)$ when $X \sim N(220, 20^2)$.

### Example 2: Compute $P(X \geq 260)$ when $X \sim N(220, 20^2)$.

# Part 3: Central Limit Theorem / Normal Approximation

Central limit theorem says the sample mean of $n$ i.i.d. (independent
and identically distributed) random variables,
$\overline{X}=\frac{1}{n}\sum_{i=1}^n X_i$, is approximately Normally
distributed, with mean $\mu=E[X]$ and variance
$\sigma^2=Var(\overline{X}) = Var(X)/n$. In other words,
$$\frac{\overline{X}-E[X]}{\sqrt{Var(\overline{X})}}\approx N(0,1).$$

It is important to note that while $E[\overline{X}]=E[X]$ (by linearity
of expectation and identical mean for each $X_i$), the variance of the
sample mean has an additional factor of $1/n$:
$$Var(\overline{X})=Var(X)/n.$$

A special case that you need to remember: When $X_i$ are Bernoulli, we
can remember the same result in the sum format (instead of the mean):

$\text{Binomial}(n, p)$ is approximately Normal with mean $\mu=np$ and
s.d. $\sigma = \sqrt{np(1-p)}$.

## An illustration

```{r clt-binomial, fig.width=10, fig.height=8}
set.seed(42)
B <- 5000
p  <- 0.4
ns <- c(10, 20, 50, 100)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (n in ns) {
  means <- replicate(B, mean(rbinom(n, size = 1, prob = p)))
  
  mu  <- p
  sig <- sqrt(p * (1 - p) / n)
  
  hist(means,
       breaks  = 20,
       probability = TRUE,
       col     = "steelblue",
       border  = "white",
       xlim    = c(mu - 4*sig, mu + 4*sig),
       ylim    = c(0, dnorm(0, sd = sig) * 1.2),
       main    = paste0("Binomial(n = ", n,", p = ",p, ")"),
       xlab    = "Sample mean",
       ylab    = "Density")
  
  curve(dnorm(x, mean = mu, sd = sig),
        col = "tomato", lwd = 2.5, add = TRUE)
}
```

## CLT simulation from a skewed population

We simulate from an exponential distribution. The original distribution
is right-skewed, definitely not normal. But the sample mean becomes more
and more normal-looking as the sample size increases.

```{r}
set.seed(42)
B <- 5000

means_n1  <- replicate(B, mean(rexp(1,  rate = 1)))
means_n5  <- replicate(B, mean(rexp(5,  rate = 1)))
means_n30 <- replicate(B, mean(rexp(30, rate = 1)))

par(mfrow = c(1, 3))

hist(means_n1, 
     breaks = 30, 
     probability = TRUE,
     main = "Sample mean, n = 1",
     xlab = "xbar", col = "steelblue", border = "white")
curve(dnorm(x, mean = 1, sd = 1/sqrt(1)), 
      add = TRUE, col = "tomato", lwd = 2)

hist(means_n5, 
     breaks = 30, 
     probability = TRUE,
     main = "Sample mean, n = 5",
     xlab = "xbar", col = "steelblue", border = "white")
curve(dnorm(x, mean = 1, sd = 1/sqrt(5)),
      add = TRUE, col = "tomato", lwd = 2)

hist(means_n30, 
     breaks = 30, 
     probability = TRUE,
     main = "Sample mean, n = 30",
     xlab = "xbar", col = "steelblue", border = "white")
curve(dnorm(x, mean = 1, sd = 1/sqrt(30)), 
      add = TRUE, col = "tomato", lwd = 2)
```

## Example: Approximating Binomial tail probabilities using Normal approximation

**Problem.** A fair coin is flipped 100 times. Using the Normal
approximation, find the probability of obtaining at least 60 heads.
What's the probability of obtaining at least 40 and no more than 60
heads?
