# Activity 16 {#sec-week2}

Name:



1. We will first use the Boston data set, which records medv (median house value) for 506 census tracts in Boston. Use the lm function. to fit a simple linear regression model, with medv as the response and lstat (percent of households with low socioeconomic status) as the predictor. Plot the scatterplot and state your predicted model?

```{r}
library(MASS)
library(ISLR2)


```

2. You can obtain a confidence interval for the coefficient estimates with the command confint(lm.out). How do you interpret these intervals?


3. The predict function can be used to produce confidence intervals and prediction intervals for the prediction of medv for a given value of lstat. Obtain prediction intervals for lstat of 5,10 and 15. 
```{r}
#predict(lm.out,data.frame(lstat=(c(5,10,15))),interval="confidence")
```

4. Add the regression line to your plot


5. Now regress medv on lstat and age. What is your predicted model? What can you conclude?


6. Now we will try to predict per capita crime rate using the other variables ofor the same Boston data set.

(a) For each preditor, fit a simple linear regression model to predict the response. Describe your results. In which of the models is there a statistically significant association between the predictor and the response? Creaste some plots to back up your assertions.

(b) Fit a multiple regression model to predict the response using all of the predictors. Describe your results. For which predictors can we reject the null hypothesis $H_{0}: \beta_{j}=0$?


(c) How do your results from (a) compare to your results from (b)? 

(d) Is there evidene of non-linear association between any of the predictors and the response? To answer this question, for each predictor X, fit a model of the form
$Y=\beta_{0}+\beta_{1}X+\beta_{2}X^2+\beta_{3}X^3+\epsilon$

