---
title: "Week 9 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---
# Logistic Regression vs. Survival Analysis

## 1. Conceptual Goal

::: {.columns style="column-gap: 3em;"}

::: {.column width="48%"}

### Logistic Regression

Logistic regression models whether an event occurs.

Example question:

> Does the patient experience the event?
:::

::: {.column width="48%"}
### Survival Analysis

Survival analysis models when an event occurs.


Example question:

> How long until the patient experiences the event?
:::

:::


## 2. Data Structure

::: {.columns style="column-gap: 3em;"}

::: {.column width="48%"}
### Logistic Regression Data

For subject $i$, observe $(X_i, Y_i)$,

where $Y_i =
\begin{cases}
1, & \text{event occurs} \\
0, & \text{event does not occur}
\end{cases}$

Time is ignored and censoring is not handled.

Example:

| patient | treatment | age | event |
|---|---:|---:|---:|
| 1 | 1 | 60 | 1 |
| 2 | 0 | 55 | 0 |
:::

::: {.column width="48%"}
### Survival Analysis Data

For subject $i$, observe $(X_i, Y_i, \delta_i)$,

where $Y_i = \min(T_i, C_i)$

and $\delta_i =\begin{cases}
1, & T_i \le C_i \quad \text{(event observed)} \\
0, & T_i > C_i \quad \text{(right censored)}
\end{cases}$


Time and censoring is directly modeled.

Example:

| patient | treatment | age | time | status |
|---|---:|---:|---:|---:|
| 1 | 1 | 60 | 4.2 | 1 |
| 2 | 0 | 55 | 6.0 | 0 |
:::

:::


## 3. Regression Model Form

::: {.columns style="column-gap: 3em;"}

::: {.column width="48%"}

### Logistic Regression

<div style="
display: inline-block;
border: 1px solid #d0d0d0;
border-radius: 8px;
padding: 10px 16px;
margin-top: 8px;
margin-bottom: 8px;
">

Log-odds form:

$\log \frac{P(Y=1 \mid x_1, x_2)}
{1-P(Y=1 \mid x_1, x_2)}
=
\beta_0 + \beta_1 x_1 + \beta_2 x_2$
</div>

<div style="color: gray;font-size: 0.9em;">

Odds form (intermediate step):

$\frac{P(Y=1 \mid x_1, x_2)}
{1-P(Y=1 \mid x_1, x_2)}
=
\exp(\beta_0 + \beta_1 x_1 + \beta_2 x_2)$
</div>

<div style="
display: inline-block;
border: 1px solid #d0d0d0;
border-radius: 8px;
padding: 10px 16px;
margin-top: 8px;
margin-bottom: 8px;
">

Odds ratio form:

$\frac{
\text{Odds}(Y=1 \mid x_1+1, x_2)
}{
\text{Odds}(Y=1 \mid x_1, x_2)
}
=
\exp(\beta_1)$
</div>

where

- $\text{Odds}(Y=1 \mid x_1, x_2)=\frac{P(Y=1 \mid x_1, x_2)}{1-P(Y=1 \mid x_1, x_2)}$ is the odds of the event occurring

:::

::: {.column width="48%"}

### Cox Survival Model

<div style="
display: inline-block;
border: 1px solid #d0d0d0;
border-radius: 8px;
padding: 10px 16px;
margin-top: 8px;
margin-bottom: 8px;
">

Log-hazard form:

$\log h(t \mid x_1, x_2)
=
\log h_0(t)
+
\beta_1 x_1 + \beta_2 x_2$
</div>

<div style="color: gray;font-size: 0.9em;">

Hazard form (intermediate step):

$h(t \mid x_1, x_2)
=
h_0(t)\exp(\beta_1 x_1 + \beta_2 x_2)$
</div>

<div style="
display: inline-block;
border: 1px solid #d0d0d0;
border-radius: 8px;
padding: 10px 16px;
margin-top: 8px;
margin-bottom: 8px;
">

Hazard ratio form:

$\frac{
h(t \mid x_1+1, x_2)
}{
h(t \mid x_1, x_2)
}
=
\exp(\beta_1)$
</div>

where

- $h(t \mid x_1,x_2)$ is the hazard at time $t$

- $h_0(t)$ is the baseline hazard

:::

:::

## 4. Meaning of the Coefficients

::: {.columns style="column-gap: 3em;"}

::: {.column width="48%"}
### Logistic Regression

- Log-odds interpretation: a one-unit increase in $X_j$ changes the log-odds of the event by $\beta_j$.

- Odds ratio interpretation: $\exp(\beta_j)$ is the multiplicative change in the odds of the event.
  - If $\beta_j = 0$, then $\exp(\beta_j)=1$, no change in odds.
  - If $\beta_j > 0$, then $\exp(\beta_j)>1$, higher odds of the event.
  - If $\beta_j < 0$, then $\exp(\beta_j)<1$, lower odds of the event.
:::

::: {.column width="48%"}
### Survival Analysis

- Log-hazard interpretation: a one-unit increase in $X_j$ changes the log-hazard of the event by $\beta_j$.

- Hazard ratio interpretation: $\exp(\beta_j)$ is the multiplicative change in the instantaneous event risk.
  - If $\beta_j = 0$, then $\exp(\beta_j)=1$, no change in hazard.
  - If $\beta_j > 0$, then $\exp(\beta_j)>1$, higher event risk and typically shorter survival.
  - If $\beta_j < 0$, then $\exp(\beta_j)<1$, lower event risk and typically longer survival.
:::

:::


## 5. Toy Example: Regression Tables

### Logistic Regression

```{r}
#| echo: false

set.seed(1)

n <- 200

x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)

eta <- -0.5 + 1.2 * x1 - 0.8 * x2

p <- 1 / (1 + exp(-eta))

y <- rbinom(n, 1, p)

dat_logistic <- data.frame(y, x1, x2)

fit_logistic <- glm(
  y ~ x1 + x2,
  data = dat_logistic,
  family = binomial
)

summary(fit_logistic)
```

---

### Cox Survival Model

```{r}
#| echo: false

library(survival)

set.seed(1)

n <- 200

x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.5)

lambda0 <- 0.1

rate <- lambda0 * exp(1.0 * x1 - 0.7 * x2)

event_time <- rexp(n, rate = rate)
censor_time <- rexp(n, rate = 0.05)

time <- pmin(event_time, censor_time)

status <- as.numeric(event_time <= censor_time)

dat_surv <- data.frame(
  time,
  status,
  x1,
  x2
)

fit_cox <- coxph(
  Surv(time, status) ~ x1 + x2,
  data = dat_surv
)

summary(fit_cox)
```

## 6. Interpretation of Toy Results

::: {.columns style="column-gap: 3em;"}

::: {.column width="48%"}
### Logistic Regression

- *Log-odds* interpretation: a one-unit increase in $x_1$ increases the log-odds of the event by $\beta_1 \approx 1.118$.

- *Odds ratio* interpretation: $\beta_2 < 0 \Leftrightarrow \exp(\beta_2) < 1$, then subjects with higher $x_2$ have lower odds of the event.
:::

::: {.column width="48%"}
### Cox Survival Model

- *Log-hazard* interpretation: a one-unit increase in $x_1$ increases the log-hazard of the event by $\beta_1 \approx 1.116$.

- *Hazard ratio* interpretation: $\beta_2 < 0 \Leftrightarrow \exp(\beta_2) < 1$, then subjects with higher $x_2$ have lower instantaneous event risk and therefore tend to survive longer.
:::

:::


## 7. Summary

| Aspect | Logistic Regression | Survival Analysis |
|---|---|---|
| Outcome | Binary event | Time-to-event |
| Data | $(X_i, Y_i)$ | $(X_i, Y_i, \delta_i)$ |
| Model target | Event probability | Hazard over time |
| Main coefficient scale | Odds ratio | Hazard ratio |
| Uses event time? | No | Yes |
| Handles censoring? | No | Yes |
