# Activity 10 {#sec-week2}

Name:

1.  The following table shows the number of bacteria colonies present in each of several petri dishes, after E. coli bacteria were added to the dishes and they were incubated for 24 hours. The “soap” dishes contained a solution prepared from ordinary soap; the “control” dishes contained a solution of sterile water. Data for the control group is: 30, 36, 66 and for the soap group: 76, 27, 16. There are 20 possible randomizations of the six observations into two groups, each of size three.

    a) Create a list of these 20 randomizations (one of which is the original assignment of observations to the two groups) and for each case calculate the control mean minus the soap mean. How many of the 20 randomizations produce a difference in means at least as large as 4.3?

```{r}
x<-c(30, 36, 66, 76, 27, 16) 
idxC <- combn(1:6, 3)

```
    b) Is there evidence that the soap inhibits E. coli growth? Justify your answer using the randomization results.


2. Suppose we have conducted a test, with $\alpha = 0.10$,
and the p-value is 0.07. For each of the following statements, say whether the statement is true or false and explain why.
(a) We reject with $H_0$ with $\alpha=0.10$
(b) We have significant evidence for $H_{A}$ with $\alpha=0.1$
(c) We would reject $H_{0}$ if were 0.05.
(d) We do not have significant evidence for $H_{A}$ with $\alpha=0.05$
(e) The probability that $\bar{Y}_{1}$ is greater than $\bar{Y}_{2}$ is 0.07.


3. Substances to be tested for cancer-causing poten-
tial are often painted on the skin of mice. The question
arose whether mice might get an additional dose of the
substance by licking or biting their cagemates. To answer
this question, the compound benzo(a)pyrene was applied
to the backs of 10 mice: Five were individually housed
and 5 were group-housed in a single cage. After 48 hours,
the concentration of the compound in the stomach tissue
of each mouse was determined. The results (nmol/gm)
were as follows:

```{r}
single<-c(3.3,2.4,2.5,3.3,2.4)
group<-c(3.9,4.1,4.8,3.9,3.4)

```

Use a Wilcoxon-Mann-Whitney test to compare the
two distributions at $\alpha = 0.01$. Let the alternative 
hypothesis be that benzo(a)pyrene concentrations
tend to be higher in group-housed mice than in
singly housed mice.

4. In an experiment to compare two diets for fattening beef steers, nine pairs of animals were chosen from the herd; members of each pair were matched as closely as possible with respect to hereditary factors. The members of each pair were randomly allocated, one to each diet. The following data shows the weight gains (lb) of the animals over a 140-day test period on diet 1 (Y1) and
on diet 2 (Y2).

```{r}
diet1<-c(596,422,524,454,538,552,478,564,556)
diet2<-c(498,460,468,458,530,482,528,598,456)
```

(a) Calculate the standard error of the mean difference.

(b) Test for a difference between the diets using a paired t-test at
$\alpha = 0.10$. Use a nondirectional alternative.

(c) Construct a 90% confidence interval for $\mu_D$.

(d) Interpret the confidence interval from part (c) in the
context of this setting.
