---
#title: "Week 3 Section"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
resources:
  - Week3_section_download.qmd
---

```{=html}
<style>
.header-row {
  display: flex;
  justify-content: space-between; /* title left, button right */
  align-items: center;
  margin-bottom: 1em; /* spacing below header */
}

.page-title {
  font-size: 2rem;   /* matches H1 style */
  font-weight: 700;
  margin: 0;
}

.download-btn {
  display: inline-block;
  background-color: #5FA2F3;  /* lighter R blue */
  color: white !important;
  padding: 8px 16px;
  border-radius: 8px;
  text-decoration: none;
  font-weight: 600;
  font-size: 14px;
}

.download-btn:hover {
  background-color: #4C93E6;
  color: white !important;
  text-decoration: none;
}
</style>
```

<!-- Flex container aligns title and button with text margin -->

::: header-row
<h1 class="page-title">

Week 3 Section

</h1>

<a href="Week3_section_incomplete.qmd" class="download-btn" download> ⬇
Download Starter File </a>
:::


# Part 1: Bernoulli and Binomial in Simulation

The **probability** of an event is the proportion of times it occurs in
a very large number of identical, independent repetitions of the same
random experiment. We can use random number generators in R to
*simulate* these repetitions and see this in action.

## 1. Simulating a coin flip

Suppose we flip a fair coin. The probability of getting Heads should be
0.5, i.e. $X \sim \text{Bernoulli}(0.5)$. Let's verify this by
simulation.

```{r}
# Simulate 10 coin flips: 1 = Heads, 0 = Tails

# Proportion of Heads in 10 flips

```

```{r}
# Now simulate 1,000 flips

```

The frequency gets closer and closer to the probability (0.5) as we
increase the number of flips. This illustrates the **Law of Large
Numbers**.

Alternative: using `rbinom()`. Hint:

-   Bernoulli(p): single trial (0/1 outcome, $P(X=1)=p$)

-   Binomial(n, p): sum of n independent Bernoulli(p) trials

```{r}
# 10 Bernoulli trials


# 1000 Bernoulli trials


```

### Optional: Seeds for reproducibility

```{r}
# Scenario 1:
sample(c(1, 0), size = 10, replace = TRUE)
sample(c(1, 0), size = 10, replace = TRUE)

# Scenario 2:
set.seed(123)
sample(c(1, 0), size = 10, replace = TRUE)

set.seed(123)
sample(c(1, 0), size = 10, replace = TRUE)
```

## 2. Simulating a biased coin

Now suppose the coin is biased, with $P(H) = 0.7, P(T) = 0.3$.

```{r}
# Simulate 1000 biased coin flip experiments: 1 = Heads, 0 = Tails
biased_coin <- sample(c(1, 0),
                      size = 1000,
                      replace = TRUE #,
                      # TODO: assigning sampling probability
                      )
mean(biased_coin)
```

Alternative: using `rbinom()`

```{r}
# Simulate 1000 experiments, each with one coin flip

```

The simulated proportions should be close to 0.7.

## 3. From Bernoulli to Binomial

Suppose X is the number of times that the coin landed on heads over 10
independent experiments.

### Task 1: What is the probability that $P(X \leq 6), P(X \geq 8)$?

```{r}
# Repeat the flips 10 times in each experiment


# P(X <= 6)


# P(X >= 8)

```

### Task 2: What is the 95th percentile?

```{r}
# Empirical quantile

```

### Task 3: We can also plot the empirical PMF with the following starter code. What does each bin represent?

```{r}
# Empirical PMF
# pmf_emp <- table(sim_biased) / length(sim_biased)

# barplot(pmf_emp,
#         main = "Empirical PMF of Binomial(10, 0.7)",
#         xlab = "Number of Heads",
#         ylab = "Probability",
#         col = "steelblue")
```

## 4. Compute the exact values

-   compute exact probability using `pbinom()`

-   compute exact quantile using `qbinom()`

-   compute exact PMF using `dbinom()`

### Task 1: What is the probability that $P(X \leq 6), P(X \geq 8)$?

```{r}
# Probabilities

```

### Task 2: What is the 95th percentile?

```{r}
# 95th percentile

```

### Task 3: We can also plot the exact PMF with the following starter code. How does it compare with the empirical PMF?

```{r}
# Values of X (number of successes)
k <- 0:10

# Exact PMF for Binomial(n = 10, p = 0.7)
pmf_exact <- dbinom(k, size = 10, prob = 0.7)

# Plot
barplot(pmf_exact,
        names.arg = k,
        main = "Exact PMF of Binomial(10, 0.7)",
        xlab = "Number of Heads",
        ylab = "Probability",
        col = "orange")
```

------------------------------------------------------------------------

# Part 2: Work by Hand

## 5. Compute the following exact probabilities:

Let $X \sim \text{Bernoulli}(p)$, where $P(X = 1) = 0.7$,
$P(X = 0) = 0.3$.

-   $P(X=7)$

-   $P(X\leq 6)$

-   $P(X\geq 8)$

Hint:

-   $P(X = k) = \binom{n}{k} p^k (1 - p)^{n - k}$

-   Use symmetry or complements when helpful

## 6. Compute mean for Bernoulli and Binomial.

(a) Bernoulli Random Variable

Let $X \sim \text{Bernoulli}(p)$, where $P(X = 1) = p$,
$P(X = 0) = 1-p$, $p=0.7$, compute the mean.

(b) Binomial Random Variable

Let $X \sim \text{Binomial}(n, p)$, $n=1000$, $p=0.7$, compute the mean.
Hint: we can write $X=X_1+X_2+ \cdots+X_n$, where each
$X_i\sim \text{Bernoulli}(p)$.

## Extra practice: Simulating a die

### Calculating probability of an event: rolling two dice

Now let's simulate rolling two fair six-sided dice and ask: **What is
the probability that the sum equals 7? What about 8?**

```{r}
n_rolls <- 10000   # number of times we roll the pair of dice

die1 <- sample(1:6, size = n_rolls, replace = TRUE)
die2 <- sample(1:6, size = n_rolls, replace = TRUE)

sums <- die1 + die2

# Estimated probability that sum = 7

# Estimated probability that sum = 8

```

We can compare to the **exact** probabilities. There are
$6 \times 6 = 36$ equally likely outcomes. The ways to get a sum of 7
are: (1,6), (2,5), (3,4), (4,3), (5,2), (6,1) — that is 6 outcomes. For
sum = 8: (2,6), (3,5), (4,4), (5,3), (6,2) — that is 5 outcomes.

```{r}
# Exact probabilities

```

Our simulated values should be very close! (Because we simulated many
identical, independent repetitions of this experiment.)

We can also look at the full distribution of the sum of two dice:

```{r}
# Frequency table of all possible sums
table(sums) / n_rolls
```

```{r}
# Visualize
barplot(table(sums) / n_rolls,
        main = "Simulated Probability Distribution of Sum of Two Dice",
        xlab = "Sum of Two Dice",
        ylab = "Estimated Probability",
        col = "steelblue")

# Add a horizontal line at the exact P(sum=7) = 6/36 for reference
abline(h = 6/36, col = "red", lty = 2)
```

Notice the distribution is **symmetric** around 7 and **triangular** in
shape — sums close to 7 are most likely.

### Simulate a biased die

Now let the die be biased. Suppose the probabilities are:

-   1 with probability 0.05
-   2 with probability 0.10
-   3 with probability 0.15
-   4 with probability 0.20
-   5 with probability 0.20
-   6 with probability 0.30

```{r}
prob_die <- c(0.05, 0.10, 0.15, 0.20, 0.20, 0.30)
sum(prob_die)
```

```{r}
biased_rolls <- sample(1:6, 
                       size = 10000,
                       replace = TRUE#,
                       # TODO: assigning sampling probability
                       )
head(biased_rolls)
```

### What is the probability of rolling a 5 or 6?

```{r}

```

Exact answer from the probability distribution:

```{r}

```

### Display the distribution of the biased die

Because the outcomes are discrete, a barplot is often the first choice.

```{r}
barplot(table(biased_rolls),
        main = "Biased die outcomes",
        xlab = "Outcome",
        ylab = "Count",
        col = "tomato")
```

We can also draw a histogram. `hist(biased_rolls, ...)`

### Sum of two rolls of the biased die

We can also look at the distribution of the sum of two biased dice.
Following code is essentially same as before, this time with
`prob = prob_die` in `sample()`.

```{r}
n_rolls <- 10000   # number of times we roll the pair of dice

die1 <- sample(1:6, size = n_rolls, replace = TRUE, prob = prob_die)
die2 <- sample(1:6, size = n_rolls, replace = TRUE, prob = prob_die)

sums_biased <- die1 + die2

barplot(table(sums_biased) / n_rolls,
        main = "Simulated Probability Distribution of Sum of Two Dice",
        xlab = "Sum of Two Dice",
        ylab = "Estimated Probability",
        col = "tomato")
```

### Compare it with the same for the fair dice

Reproducing the plot from earlier (no need to run the experiment
again!):

```{r}
barplot(table(sums) / n_rolls,
        main = "Simulated Probability Distribution of Sum of Two Dice",
        xlab = "Sum of Two Dice",
        ylab = "Estimated Probability",
        col = "steelblue")
```
